---
title: "Bio_sample_Subclass"
output: html_document
date: "2026-04-20"
---

```{r}
library(Spectra)
library(RuSirius)
library(MsCoreUtils)
library(tidyverse)
```

Since ww are working on untargeted dataset, the chemciual classification of the detected features is not known.
The plan is to forecast the chemical subclass for each feature using Sirius softwer.
RuSirius pakage can be used to communicate with SIRIUS througth R environment.
Fist, we load the MSMS data from the precedent step 'Biological_sample_workflow', so we have negative, positive and neutral spectra of considered features 

```{r}
# loading MSMS data
load("rdata/alt_msms.RData")
```

Now for each original polarities (positive and negative MSMS) we run a SIRIUS experiment in order to ave the prediction of chemicla classes for each features
We assume all feature considered are [M+H]+ and [M-H]- adducts.

POSITIVE ESI

```{r}
# PS: opening SIRIUS with a registered account
srs <- Sirius()

#selecting spectra
sp_subset <- MS2pos
sp_subset <- setBackend(sp_subset, MsBackendMemory())
sp_subset <- filterEmptySpectra(sp_subset)

srs <- import(sirius = srs,
              spectra = sp_subset,
              ms_column_name = "feature_id",
              deleteExistingFeatures = TRUE)

#run Sirius analysis

run(srs,
    fallbackAdducts = c("[M + H]+"),
    formulaIdParams = formulaIdParam(numberOfCandidates = 10,
                                       instrument = "ORBITRAP",
                        numberOfCandidatesPerIonization = 3,
                        massAccuracyMS2ppm = 5,
                        filterByIsotopePattern = FALSE,
                        isotopeMs2Settings = c("SCORE"),
                        performDeNovoBelowMz = 600,
                        minPeaksToInjectSpecLibMatch = 3),
    predictParams = predictParam(),

    structureDbSearchParams = structureDbSearchParam(
          structureSearchDbs = c("BIO")
      ),
    recompute = TRUE,
    wait = TRUE
    )
srs_pos <- srs
res_pos <- summary(sirius = srs_pos, result.type = "structure")
class_pos <- results(srs_pos, result.type="compoundClass",
                     return.type = "data.frame")

```

NEGATIVE ESI

```{r}
# PS: opening SIRIUS with a registered account
srs <- Sirius()

#selecting spectra
sp_subset <- MS2neg
sp_subset <- setBackend(sp_subset, MsBackendMemory())
sp_subset <- filterEmptySpectra(sp_subset)

srs <- import(sirius = srs,
              spectra = sp_subset,
              ms_column_name = "feature_id",
              deleteExistingFeatures = TRUE)

#run Sirius analysis

run(srs,
    fallbackAdducts = c("[M - H]-"),
    formulaIdParams = formulaIdParam(numberOfCandidates = 10,
                                       instrument = "ORBITRAP",
                        numberOfCandidatesPerIonization = 3,
                        massAccuracyMS2ppm = 5,
                        filterByIsotopePattern = FALSE,
                        isotopeMs2Settings = c("SCORE"),
                        performDeNovoBelowMz = 600,
                        minPeaksToInjectSpecLibMatch = 3),
    predictParams = predictParam(),

    structureDbSearchParams = structureDbSearchParam(
          structureSearchDbs = c("BIO")
      ),
    recompute = TRUE,
    wait = TRUE
    )
srs_neg <- srs
res_neg<- summary(sirius = srs_neg, result.type = "structure")
class_neg <- results(srs_neg, result.type="compoundClass",
                     return.type = "data.frame")

```

Now, for each feature, we have chemical classification for both ESI.
For each feature we extract the subclass with the highest probability.
We consider a reliable prediction only the once with the same subclass prediction in both ESI.

```{r}
#loop to identify one class and subclass for each feature
#we store the results in the 'comonft' dataframe, which has been used for neutral pseudo-spectra pairing
comonft$class_neg <- NA
comonft$subclass_neg <- NA
comonft$class_pos <- NA
comonft$subclass_pos <- NA

for (i in 1:nrow(comonft)){
  #negative part
   n <- comonft$name_n[i]
   sb <- class_neg[class_neg$xcms_fts== n,] %>%
     filter(type=="ClassyFire") 
   
   if(nrow(sb)>0){
     cl <- sb %>%
       filter(level =="Class") %>%
       slice_max(probability, n=1) %>%
       pull(name)
    if(length(cl)>0){
      comonft$class_neg[i] <- cl
    }
      scl <- sb %>%
       filter(level =="Subclass") %>%
       slice_max(probability, n=1) %>%
       pull(name)
 if(length(scl)>0){
        comonft$subclass_neg[i] <- scl
 }
   }
   
    #positive part
   p <- comonft$name_p[i]
   sb <- class_pos[class_pos$xcms_fts== p,] %>%
     filter(type=="ClassyFire") 
   
   if(nrow(sb)>0){
     cl <- sb %>%
       filter(level =="Class") %>%
       slice_max(probability, n=1) %>%
       pull(name)
      if(length(cl)>0){
        comonft$class_pos[i] <- cl
        }
      scl <- sb %>%
       filter(level =="Subclass") %>%
       slice_max(probability, n=1) %>%
       pull(name)
          if(length(scl)>0){
        comonft$subclass_pos[i] <- scl
        }
   }
 }


#consensus classes 
comonft$sbclass_neu  <- NA
comonft$class_neu  <- NA

comonft$sbclass_neu <- ifelse(comonft$subclass_neg == comonft$subclass_pos, comonft$subclass_pos, NA) 
comonft$class_neu <- ifelse(comonft$class_neg == comonft$class_pos, comonft$class_pos, NA) 

```

Now we evaluate the networks performance with modularity mesurements based on subclass labelling
Different networks can be build considering different MSMS form the same feature (compound). For this, initial xcms data from both polarities are needed.
The aim is to initially create a id for each MSMS detected in initial DDA.

```{r}
#extract all msms realted to features (loaded during xcms processing)

#positive msms
load("rdata/alt_pos_xcms.RData")

msms <- Spectra(sampleData(mse)$spectraOrigin)
msms$idf <- paste0(msms$scanIndex, basename(msms$dataOrigin),round(msms$precursorIntensity))
msms$feature_id <- NA

for(x in 1:length(MS2pos)){
  mzx <-precursorMz(MS2pos)[x]
  rtx <- rtime(MS2pos)[x]
  
  msms$feature_id[which(MsCoreUtils::between(msms$precursorMz, mzx +0.001*c(-1,1))&
          MsCoreUtils::between(msms$rtime, rtx+5*c(-1,1)))] <- MS2pos$feature_id[x]
}
spspos <- msms

#negative msms
load("rdata/alt_neg_xcms.RData")

msms <- Spectra(sampleData(mse)$spectraOrigin)
msms$idf <- paste0(msms$scanIndex, basename(msms$dataOrigin),round(msms$precursorIntensity))
msms$feature_id <- NA

for(x in 1:length(MS2neg)){
  mzx <-precursorMz(MS2neg)[x]
  rtx <- rtime(MS2neg)[x]
  
  msms$feature_id[which(MsCoreUtils::between(msms$precursorMz, mzx +0.001*c(-1,1))&
          MsCoreUtils::between(msms$rtime, rtx+5*c(-1,1)))] <- MS2neg$feature_id[x]
}
spsneg<- msms

metadata_pos <- as.data.frame(spectraData(spspos))
metadata_neg <- as.data.frame(spectraData(spsneg))
```

Now we calculate the modularity (global) and also local modularity (for given chemical subclasses) across different similarity thresholds (0-1).
In each repetition, for each feature, a related MSMS is randomly chosen between the avaiable one, for each polarity. Neutral pseudo-spectra are then re-buit and similarity matrix re-calculated. Note this can be a computationally heavy task.

```{r}
#utility functios

    #to evaluate connectivity of given names and overall modularity
      eval_net <- function(adj_matrix, node_names_ref, cm2_names_ref) {
          adj <- adj_matrix
          adj[adj < threshold] <- 0
          diag(adj) <- 0
          g <- graph_from_adjacency_matrix(adj, mode = "undirected", weighted = TRUE, diag = FALSE)
          
          V(g)$class <- comonft$class[match(V(g)$name, node_names_ref)]
          V(g)$subclass <- comonft$subclass[match(V(g)$name, node_names_ref)]
          
          tot <- numeric(nrow(df)); corr <- numeric(nrow(df)); wrong <- numeric(nrow(df)); mod <-        numeric(nrow(df))
          
          for(i in 1:nrow(df)) {
            node_name <- cm2_names_ref[i]
            v <- V(g)[V(g)$name == node_name]
            
            if(length(v) > 0) {
              nh <- ego(g, order = 1, nodes = v)[[1]]
              nh <- setdiff(nh, v)
              cls <- V(g)$subclass[nh]
              or_cl <- cm2$Subclass[which(cm2_names_ref == node_name)]
              
              tot[i] <- length(cls)
              corr[i] <- length(which(cls == or_cl))
              wrong[i] <- length(which(cls != or_cl & !is.na(cls)))
            }
          }
          
          mem <- as.numeric(factor(V(g)$subclass))
          mem[is.na(mem)] <- 99
          mod <- modularity(g, membership = mem)
          
          return(list(tot = tot, corr = corr, wrong = wrong, mod = mod))
      }

      #function to handle overlapping ions
      process_data <- function(data) {
        data <- data %>%
          arrange(mz) %>%
          mutate(group = cumsum(c(TRUE, diff(mz) >= 0.001))) %>%
          group_by(group) %>%
          summarise(
            mz = mean(mz),
            intensity = sum(intensity)
          ) %>%
          ungroup() %>%
          dplyr::select(-group)
        return(data)
      } 
    
    #to compute LQ calacualtion
     process_lq <- function(adj_mat, mode_name, node_names_ref) {
          adj <- adj_mat
          adj[adj < threshold] <- 0
          diag(adj) <- 0
          g <- graph_from_adjacency_matrix(adj, mode = "undirected", weighted = TRUE, diag = FALSE)
          V(g)$subclass <- comonft$subclass[match(V(g)$name, node_names_ref)]
          
          res <- sapply(target_cls, function(x) {
            nodes_in_cls <- V(g)$name[which(V(g)$subclass == x)]
            if(length(nodes_in_cls) > 1) {
              tryCatch(LQ(g, nodes_in_cls), error = function(e) NA)
            } else { NA }
          })
          return(data.frame(thr = threshold, Mode = mode_name, Class = target_cls, LQ = res))
     }
     
# loop for MODs calculation across different thresholds

target_cls <- c("Sesquiterpenoids", "Amino acids, peptides, and analogues", "Terpene lactones", "Diterpenoids")  #for LQ calculation

valtab <- data.frame(rep=1:25)
valtab$data <- vector(mode="list", length=nrow(valtab))
valtab$LQ <- vector(mode="list", length=nrow(valtab))

# OUTER LOOP: msms boostrap

for(rep_idx in 1:nrow(valtab)) {
  cat(sprintf("-> Execution %d of %d\n", rep_idx, nrow(valtab)))
  
  #positive selection
  unique_ids <- metadata_pos %>%
    group_by(feature_id) %>%
    slice_sample(n = 1) %>%      #random sampling one msms for each feature
    ungroup() %>%
    filter(feature_id %in% MS2neu$feature_id)
  tmppos<- spspos[match(unique_ids$idf, spspos$idf)] %>%
    cleaning_ref()
  
  #neagtive selection
  unique_ids <- metadata_neg %>%
    group_by(feature_id) %>%
    slice_sample(n = 1) %>%      #random sampling one msms for each feature
    ungroup() %>%
    filter(feature_id %in% MS2neg$feature_id)
  tmpneg<- spsneg[match(unique_ids$idf, spsneg$idf)] %>%
    cleaning_ref()
  
  #neutral formation
  nn<- comonft$name_p[match(tmpneg$feature_id,comonft$name_n)]
  pn <- unique(nn, tmppos$feature_id)
  
  meta <- comonft %>%
    dplyr::select(rtmed, name_p, name_n) 
  meta$feature_id <- meta$name_p
  
  meta$precursorMz <- NA
  meta$mz <- meta$intensity <- vector(mode="list", length=nrow(meta))
  
  for(i in 1:nrow(meta)){
    register(SerialParam())
    
    #positive
    akap <- tmppos[which(tmppos$feature_id == meta$feature_id[i])]
    if(length(akap)>0){
      mzpos <- mz(akap)[[1]] 
      mzpos <- mzpos - 1.00728
      intpos <- intensity(akap)[[1]]
      intpos <- (intpos/max(intpos))*100
      df_pos <- data.frame(mz = mzpos, intensity = intpos)
      
      #negative
      akan <- tmpneg[which(tmpneg$feature_id == meta$name_n[i])]
      if(length(akan)>0){
        mzneg <- mz(akan)[[1]] 
        mzneg <- mzneg + 1.00728
        intneg <- intensity(akan)[[1]]
        intneg <- (intneg/max(intneg))*100
        df_neg <- data.frame(mz = mzneg, intensity = intneg) 
        
        #both
        df <- bind_rows(df_neg, df_pos) %>%
          arrange(mz)
        df<- as.data.frame(process_data(df))
        meta$mz[[i]]<- as.numeric(df$mz)
        meta$intensity[[i]] <- as.numeric(df$intensity)
        
        meta$precursorMz[i] <- mean(c(precursorMz(akan),precursorMz(akap)))
      }
    }
    meta_all <- rbind(meta, meta_all)
  }
  
  tmpneu <- Spectra(meta_all)
  
  #create matrices
  register(SerialParam())
  mat_neu <- compareSpectra(tmpneu, MAPFUN = joinPeaksGnps, tolerance = 0.01, FUN = MsCoreUtils::gnps)
  colnames(mat_neu)<- rownames(mat_neu) <- tmpneu$feature_id
  mat_neg <- compareSpectra(tmpneg, MAPFUN = joinPeaksGnps, tolerance = 0.01, FUN = MsCoreUtils::gnps)
  colnames(mat_neg)<- rownames(mat_neg) <- tmpneg$feature_id
  mat_pos <- compareSpectra(tmppos, MAPFUN = joinPeaksGnps, tolerance = 0.01, FUN = MsCoreUtils::gnps)
  colnames(mat_pos)<- rownames(mat_pos) <- tmppos$feature_id
  
  #results of MOD and LQ
  seq <- seq(0,1,0.05)
  
  main <- data.frame(thr = seq, mod_neu = NA, mod_pos = NA, mod_neg = NA)
  main$data <- vector("list", length(seq))
  lq_results <- list()
  
  #INNER LOOP: Similarity thr
  
  for(k in 1:nrow(main)){
    threshold <- main$thr[k]
    
    # D.1 LQ CALCULATION
    
    # Run for all three modes
    res_neu <- process_lq(mat_neu, "Neutral", comonft$name_p)
    res_pos <- process_lq(mat_pos, "Positive", comonft$name_p)
    res_neg <- process_lq(mat_neg, "Negative", comonft$name_n)
    
    lq_results[[k]] <- bind_rows(res_neu, res_pos, res_neg)
    
    # D.2 CONNECTIVITY & MODULARITY
    df <- data.frame(
      id = cm2$name_p[which(!is.na(cm2$Subclass))],
      tot_conn_neu = NA, corr_conn_neu = NA, wrong_conn_neu = NA,
      tot_conn_pos = NA, corr_conn_pos = NA, wrong_conn_pos = NA,
      tot_conn_neg = NA, corr_conn_neg = NA, wrong_conn_neg = NA
    )
    
    #repeat for the 3 polarity
    net_neu <- eval_net(mat_neu, comonft$name_p, cm2$name_p)
    df$tot_conn_neu <- net_neu$tot; df$corr_conn_neu <- net_neu$corr; df$wrong_conn_neu <- net_neu$wrong
    main$mod_neu[k] <- net_neu$mod
    
    net_pos <- eval_net(mat_pos, comonft$name_p, cm2$name_p)
    df$tot_conn_pos <- net_pos$tot; df$corr_conn_pos <- net_pos$corr; df$wrong_conn_pos <- net_pos$wrong
    main$mod_pos[k] <- net_pos$mod
    
    net_neg <- eval_net(mat_neg, comonft$name_n, cm2$name_n)
    df$tot_conn_neg <- net_neg$tot; df$corr_conn_neg <- net_neg$corr; df$wrong_conn_neg <- net_neg$wrong
    main$mod_neg[k] <- net_neg$mod
    
    main$data[[k]] <- df
  }
  
  valtab$data[[rep_idx]] <- main
  valtab$LQ[[rep_idx]] <- bind_rows(lq_results)
}
```









