---
title: "merging_poneg"
output: html_document
date: "2024-06-27"
---
```{r}
library(xcms)
library(dplyr)
library(tidyr)
library(purrr)
library(MsCoreUtils)
library(Spectra)
library(MsBackendMgf)
library(MetaboCoreUtils)
library(igraph)
library(visNetwork)
library(MsExperiment)
library(RuSirius)
library(MetaboAnnotation)
library(RColorBrewer)
library(MASS)
library(MsFeatures)
library(ggplot2)


cleaning_ref<- function(ms2){
  ms2_tb <- as.data.frame(
    spectraData(ms2) 
  ) %>%
  mutate(ms2_raw = map2(mz(ms2), intensity(ms2), ~{
    if (length(.x) == 0 || length(.y) == 0)
      return(data.frame(mz = numeric(0), intensity = numeric(0)))
    data.frame(mz = as.numeric(.x), intensity = as.numeric(.y))
  })) %>%
    mutate(ms2_RLM = purrr::map(ms2_raw, function(t){
         if (is.null(t) || nrow(t) == 0 || !"intensity" %in% names(t))
        return(tibble(mz = numeric(0), intensity = numeric(0)))
      t <- t %>%
        filter(intensity>0) %>%
        arrange(intensity) %>%
        mutate(order = seq(nrow(.)))
      p75 <- t %>%
        pull(intensity) %>% 
        quantile(0.75)
      u <- t %>%
        filter(intensity < p75)
      if(nrow(u)>3){
        mdl <- rlm(data = u, log10(intensity) ~ order)
        t_filt<- t %>% 
          mutate(pred = predict(mdl, t)) %>%
          mutate(res = abs(pred - log10(intensity))) %>%
          filter(res > sd(residuals(mdl))*3) %>%
          arrange(mz)
        if (nrow(t_filt) > 0) {
          t<- t_filt
        } else {
          t<- t %>% arrange(mz)
        }
      } else {
        return(t %>% arrange(mz))  # Not enough points to model
      }
    }))
  ms2_cl <- ms2_tb %>%
    mutate(
      mz = purrr::map(ms2_RLM, ~ .x$mz),
      intensity = purrr::map(ms2_RLM, ~ .x$intensity)
    ) %>%
    filter(map_lgl(mz, ~ length(.x) > 0)) %>%
    as.data.frame() %>%
    Spectra()
  return(ms2_cl)
}
```

setting parameters of RTs and mz tolerance

```{r}
rt_lim1 <- 30
rt_lim2 <- 600
tol<- 0.005 #for adducts/isopoes matching
```
we assume to have done xcms processing, which have lead to "fdata" creation for each polarities, whihc is an XCMSexperiment object created after feature creation and alignment.
First we create chromatographic groups of features, then we identify adduct within each group. [M+H]+ and [M-H]- species will be used for pairing across polarities and for neutral pseudo-spectra creation.

#GROUPING POSITIVE ESI

GROUPING

creating feature group according to rtime, correlation between sample and peak shape correlation 

```{r}
load("rdata/alt_pos_xcms.RData")

sb<- filterRt(fdata, rt=c(rt_lim1, rt_lim2))

#grouping for rtime
grdata1 <- groupFeatures(sb, param = SimilarRtimeParam(5, groupFun =groupConsecutive))
plotFeatureGroups(grdata1, pch = 21, lwd = 2, col = "#00000040", bg = "#00000020")
#for coorelation across samples
grdata2 <- groupFeatures(
    grdata1, AbundanceSimilarityParam(threshold  = 0.7, transform = log2,
                                      subset=grep("A", sampleData(grdata1)$sample)),
    filled = TRUE)
plotFeatureGroups(grdata2, pch = 21, lwd = 2, col = "#00000040", bg = "#00000020")
#for peak shape
grdata3 <- groupFeatures(grdata2, EicSimilarityParam(threshold = 0.7, n = 3))
plotFeatureGroups(grdata3, pch = 21, lwd = 2, col = "#00000040", bg = "#00000020")
```

adduct recognition for each chromatographic group

```{r}
#extracting an filtering feature table

feat_def <- as.data.frame(featureDefinitions(grdata3))
feat_def$feature_id <- rownames(feat_def)
feat_def<- feat_def[which(feat_def$feature_id %in% spspos$feature_id),]
#feat_def <- feat_def[which(feat_def$possible_contaminants !=TRUE),]

# ---------------------------------------------------------
# 1. Define Adducts and Calculate Exact Differences
# ---------------------------------------------------------
adducts <- data.frame(
  adduct = c("[M+H]+", "[M+Na]+", "[M+K]+", "[M+NH4]+", 
             "[M-H2O+H]+", "[M-2H2O+H]", "[M-CO+H]+", "[M-COOH+H]+"),
  mass = c(1.007276, 22.989218, 38.963158, 18.033823, 
           -17.003289, -34.00658, -26.987639, -44.998203)
)

# Create difference table. 
# We strictly define a1/m1 as the HEAVIER adduct to make matching foolproof.
adduct_diff <- adducts %>%
  dplyr::rename(a1 = adduct, m1 = mass) %>%
  cross_join(adducts %>%  dplyr::rename(a2 = adduct, m2 = mass)) %>%
  filter(m1 > m2) %>% # Keep only positive mass differences
  mutate(
    delta = m1 - m2,
    delta_min = delta - 0.002, # Set your tolerance here
    delta_max = delta + 0.002
  ) %>%
  filter(a1 == "[M+H]+" | a2 == "[M+H]+")

# ---------------------------------------------------------
# 2. Setup the Loop to Process Feature Groups
# ---------------------------------------------------------
results_list <- list()
grs <- unique(feat_def$feature_group)

for(gr in grs) {
  # Subset current group
  sb <- feat_def[which(feat_def$feature_group == gr), ]
  
  if(nrow(sb) > 1) {
    # Calculate distance matrix for mzmed
    mz_dists <- as.matrix(dist(sb$mzmed))
    rownames(mz_dists) <- sb$feature_id
    colnames(mz_dists) <- sb$feature_id
    
    # Extract coordinates
    idx <- which(upper.tri(mz_dists), arr.ind = TRUE)
    
    # Map back to IDs and MZ values
    feat_A <- rownames(mz_dists)[idx[, 1]]
    feat_B <- colnames(mz_dists)[idx[, 2]]
    mz_A <- sb$mzmed[match(feat_A, sb$feature_id)]
    mz_B <- sb$mzmed[match(feat_B, sb$feature_id)]
    
    pairs <- data.frame(
      f1 = ifelse(mz_A > mz_B, feat_A, feat_B),
      f2 = ifelse(mz_A > mz_B, feat_B, feat_A),
      mz_1 = ifelse(mz_A > mz_B, mz_A, mz_B), 
      mz_2 = ifelse(mz_A > mz_B, mz_B, mz_A),
      obs_delta = abs(mz_A - mz_B)
    ) %>% distinct()
    
    # Match pairs to the adduct_diff table
    matches <- pairs %>%
      inner_join(
        adduct_diff, 
        by = join_by(between(obs_delta, delta_min, delta_max))
      ) %>%
      dplyr::select(-delta_min, -delta_max)
    
    # ---------------------------------------------------------
    # 3. Build Subtable (Handling Matches AND Leftovers)
    # ---------------------------------------------------------
    if(nrow(matches) > 0) {
      # Extract annotated assignments
      sub_f1 <- matches %>% dplyr::select(feature_id = f1, adduct = a1, adduct_mass = m1)
      sub_f2 <- matches %>% dplyr::select(feature_id = f2, adduct = a2, adduct_mass = m2)
      
      annotated_subtable <- bind_rows(sub_f1, sub_f2) %>%
        distinct() %>%
        left_join(sb %>% dplyr::select(feature_id, mzmed, rtmed), by = "feature_id") %>%
        mutate(neutral_mass = mzmed - adduct_mass) %>%
        dplyr::select(feature_id, mzmed, rtmed, adduct, neutral_mass)
      
      # --- THE FIX: CATCH THE LEFTOVERS ---
      # Find features in this group that didn't get an adduct match
      leftovers <- sb %>% filter(!feature_id %in% annotated_subtable$feature_id)
      
      if(nrow(leftovers) > 0) {
        leftover_subtable <- leftovers %>%
          dplyr::select(feature_id, mzmed, rtmed) %>%
          mutate(
            adduct = "unk",
            neutral_mass = mzmed - 1.007276 # Assuming [M+H]+ for unknowns
          )
        # Bind the annotated pairs and the leftovers together
        subtable <- bind_rows(annotated_subtable, leftover_subtable) %>% arrange(mzmed)
      } else {
        subtable <- annotated_subtable %>% arrange(mzmed)
      }
      
    } else {
      # --- THE FIX: NO MATCHES FOUND AT ALL ---
      # If the group has >1 feature but no math works, treat ALL as singletons
      subtable <- sb %>%
        dplyr::select(feature_id, mzmed, rtmed) %>%
        mutate(
          adduct = "[M+H]+",
          neutral_mass = mzmed - 1.007276
        ) %>% arrange(mzmed)
    }
    
    # Save the complete group to the list
    group_rtmed <- median(subtable$rtmed)
    
    results_list[[as.character(gr)]] <- tibble(
      feature_group = gr,
      group_rtmed = group_rtmed,
      annotated_features = list(subtable),
      # This will now safely grab the M_mass of both real matches and leftovers
      M_mass = list(subtable$neutral_mass[which(subtable$adduct == "[M+H]+")]),
      n = nrow(sb)
    )
    
  } else {
    # ---------------------------------------------------------
    # TRUE SINGLETONS (Only 1 feature in the group)
    # ---------------------------------------------------------
    subtable <- tibble(
      feature_id = sb$feature_id, 
      mzmed = sb$mzmed,
      rtmed = sb$rtmed, 
      adduct = "[M+H]+",
      neutral_mass = sb$mzmed - 1.007276
    )
    
    results_list[[as.character(gr)]] <- tibble(
      feature_group = gr,
      group_rtmed = subtable$rtmed,
      annotated_features = list(subtable),
      M_mass = list(subtable$neutral_mass[which(subtable$adduct == "[M+H]+")]),
      n = 1
    )
  }
}

# ---------------------------------------------------------
# 4. Create the Final Master Table
# ---------------------------------------------------------
final_table <- bind_rows(results_list)

# ---------------------------------------------------------
# 5. refine [M+H]+ identification (handle insource fragmentation)
# ---------------------------------------------------------
loss_deltas <- c(18.0106, 36.0212, 54.0318, 27.9949, 44.9976, 43.98983)
loss_labels <- c("[M-H2O+H]+", "[M-2H2O+H]+",  "[M-3H2O+H]+", "[M-CO+H]+", "[M-COOH+H]+", "[M-CO2+H]+" )
iso_deltas <- c(1.00335, 2.00671,3.01005)

cleaned_results <- list()

for(q in 1:nrow(final_table)) {
  
  row_data <- final_table[q, ]
  sub_df <- row_data$annotated_features[[1]]
  m_candidates <- sub_df%>%
    filter(adduct=="[M+H]+") %>%
    pull(mzmed) %>%
    sort()
  
  # If there are 0 or 1 [M+H]+ candidates, keep it as is and skip
  if(length(m_candidates) < 2) {
    cleaned_results[[q]] <- row_data
    next
  }
  
  # 1. ISOTOPES
for (i in 1:(length(m_candidates)-1)) {
  for (j in (i+1):length(m_candidates)) {
    
    # Because we sorted, delta is always negative..inverte
    delta <- m_candidates[j] - m_candidates[i]
    
    if (any(abs(delta - iso_deltas) <= 0.005)) {
      true_M_iso <- m_candidates[i] 
      
      sub_df <- sub_df %>%
        mutate(
          adduct = case_when(
            # The extra "mzmed > true_M_iso" guarantees we only alter the heavier peak
            abs(mzmed - (true_M_iso + 1.00335)) <= 0.005 & mzmed > true_M_iso & adduct == "[M+H]+" ~ "[M+1+H]+",
            abs(mzmed - (true_M_iso + 2.00671)) <= 0.005 & mzmed > true_M_iso & adduct == "[M+H]+" ~ "[M+2+H]+",
            abs(mzmed - (true_M_iso + 3.01005)) <= 0.005 & mzmed > true_M_iso & adduct == "[M+H]+" ~ "[M+3+H]+",
            TRUE ~ adduct
          ),
          # Snap the neutral mass of the isotope back to the monoisotopic parent
          neutral_mass = ifelse(adduct %in% c("[M+1+H]+", "[M+2+H]+",  "[M+3+H]+"), true_M_iso, neutral_mass)
        )
    }
  }
}
  
  # 2. DIMERS
  dimer_labels <- c("[2M+H]+", "[2M+Na]+", "[2M+K]+")
  dimer_additions <- c(1.007276, 22.989218, 38.963158)
  
  for (i in 1:(length(m_candidates)-1)) {
    for (j in (i+1):length(m_candidates)) {
      
      # 1. Calculate the neutral mass of the lighter candidate (assumed monomer)
      true_M_monomer <- m_candidates[i] 
      true_neutral <- true_M_monomer - 1.007276 
      
      for (d in 1:length(dimer_labels)) {
        
        expected_dimer_mz <- (2 * true_neutral) + dimer_additions[d]
        
        if (abs(m_candidates[j] - expected_dimer_mz) <= 0.005) {
          
          target_dimer_label <- dimer_labels[d]
          sub_df <- sub_df %>%
            mutate(
              # Relabel the heavier peak to the correct dimer string
              adduct = ifelse(abs(mzmed - expected_dimer_mz) <= 0.005 & adduct == "[M+H]+", 
                              target_dimer_label, adduct),
              neutral_mass = ifelse(adduct == target_dimer_label, true_neutral, neutral_mass)
            )
        }
      } # End of dimer types loop
    }
  }
  
  # 3. INSOURCE FRAGMENTATION (Cascades)
  for (i in 1:(length(m_candidates)-1)) {
    for (j in (i+1):length(m_candidates)) {
      
      # Calculate the mass difference between the two candidates
      delta <- m_candidates[j] - m_candidates[i]
      
      # Loop through our 4 specific losses to see if there's a match
      for (k in 1:length(loss_deltas)) {
        
        if (abs(delta - loss_deltas[k]) <= 0.005) {
          fake_M_frag <- m_candidates[i]    # The lighter mass (fragment)
          true_M_cascade <- m_candidates[j] # The heavier mass (parent)
          target_label <- loss_labels[k]    # Pull the matching text label
          
          sub_df <- sub_df %>%
            mutate(
              # 1. Update the text label so we know EXACTLY what fell off
              adduct = ifelse(abs(mzmed - fake_M_frag) <= 0.005 & adduct == "[M+H]+", 
                              target_label, adduct),
                              
              # 2. Snap the neutral mass up to the true parent 
              neutral_mass = ifelse(adduct == target_label, true_M_cascade, neutral_mass)
            )
        }
      } # End of the loss checking loop
      
    }
  }
  # 4. FINALIZE THE ROW

  # Update the subtable
  row_data$annotated_features[[1]] <- sub_df
  
  # Dynamically recalculate M_mass based on the features that SURVIVED as [M-H]-
  # This perfectly preserves any unrelated singletons that were co-eluting!
  row_data$M_mass[[1]] <- unique(sub_df$neutral_mass[sub_df$adduct == "[M+H]+"])
  
  cleaned_results[[q]] <- row_data
}

# Bind it all back together into a final, pristine table
cl_table_pos<- bind_rows(cleaned_results)

ftpos<- feat_def
data_pos <- fdata
```

#GROUPING NEGATIVE ESI

creating feature group according to rtime, correlation between sample and peak shape correlation 

```{r}
load("rdata/alt_neg_xcms.RData")

sb<- filterRt(fdata, rt=c(rt_lim1, rt_lim2))

#grouping for rtime
grdata1 <- groupFeatures(sb, param = SimilarRtimeParam(5, groupFun =groupConsecutive))
plotFeatureGroups(grdata1, pch = 21, lwd = 2, col = "#00000040", bg = "#00000020")

#for correlation across samples
grdata2 <- groupFeatures(
    grdata1, AbundanceSimilarityParam(threshold  = 0.6, transform = log2,
                                      subset=grep("A", sampleData(grdata1)$sample)),
    filled = TRUE)
plotFeatureGroups(grdata2, pch = 21, lwd = 2, col = "#00000040", bg = "#00000020")
#for peak shape
grdata3 <- groupFeatures(grdata2, EicSimilarityParam(threshold = 0.7, n = 3))
plotFeatureGroups(grdata3, pch = 21, lwd = 2, col = "#00000040", bg = "#00000020")
```

adduct recognition for each chromatographic group

```{r}
#extracting an filtering feature table

feat_def <- as.data.frame(featureDefinitions(grdata3))
feat_def$feature_id <- rownames(feat_def)
feat_def<- feat_def[which(feat_def$feature_id %in% spsneg$feature_id),]
#feat_def <- feat_def[which(feat_def$possible_contaminants !=TRUE),]

# ---------------------------------------------------------
# 1. Define Adducts and Calculate Exact Differences
# ---------------------------------------------------------
adducts <- data.frame(
  adduct = c("[M-H]-", "[M+Cl]-", "[M+Br]-", "[M+HCOO]-", "[M-H2O-H]-", "[M-CO2-H]-", "[M+CH3COO]-", "[M-2H+Na]-"),
  mass = c(-1.007276,   
           34.968853,  
           78.918836,
           44.998203, 
           -19.017841,   
           -44.997105, 
           59.013853,
           20.975217)   
)


# Create difference table. 
# We strictly define a1/m1 as the HEAVIER adduct to make matching foolproof.
adduct_diff <- adducts %>%
  dplyr::rename(a1 = adduct, m1 = mass) %>%
  cross_join(adducts %>%  dplyr::rename(a2 = adduct, m2 = mass)) %>%
  filter(m1 > m2) %>% # Keep only positive mass differences
  mutate(
    delta = m1 - m2,
    delta_min = delta - 0.002, # Set your tolerance here
    delta_max = delta + 0.002
  ) %>%
  filter(a1 == "[M-H]-" | a2 == "[M-H]-")

# ---------------------------------------------------------
# 2. Setup the Loop to Process Feature Groups
# ---------------------------------------------------------
results_list <- list()
grs <- unique(feat_def$feature_group)

for(gr in grs) {
  # Subset current group
  sb <- feat_def[which(feat_def$feature_group == gr), ]
  
  if(nrow(sb) > 1) {
    # Calculate distance matrix for mzmed
    mz_dists <- as.matrix(dist(sb$mzmed))
    rownames(mz_dists) <- sb$feature_id
    colnames(mz_dists) <- sb$feature_id
    
    # Extract coordinates
    idx <- which(upper.tri(mz_dists), arr.ind = TRUE)
    
    # Map back to IDs and MZ values
    feat_A <- rownames(mz_dists)[idx[, 1]]
    feat_B <- colnames(mz_dists)[idx[, 2]]
    mz_A <- sb$mzmed[match(feat_A, sb$feature_id)]
    mz_B <- sb$mzmed[match(feat_B, sb$feature_id)]
    
    pairs <- data.frame(
      f1 = ifelse(mz_A > mz_B, feat_A, feat_B),
      f2 = ifelse(mz_A > mz_B, feat_B, feat_A),
      mz_1 = ifelse(mz_A > mz_B, mz_A, mz_B), 
      mz_2 = ifelse(mz_A > mz_B, mz_B, mz_A),
      obs_delta = abs(mz_A - mz_B)
    ) %>% distinct()
    
    # Match pairs to the adduct_diff table
    matches <- pairs %>%
      inner_join(
        adduct_diff, 
        by = join_by(between(obs_delta, delta_min, delta_max))
      ) %>%
      dplyr::select(-delta_min, -delta_max)
    
    # ---------------------------------------------------------
    # 3. Build Subtable (Handling Matches AND Leftovers)
    # ---------------------------------------------------------
    if(nrow(matches) > 0) {
      # Extract annotated assignments
      sub_f1 <- matches %>% dplyr::select(feature_id = f1, adduct = a1, adduct_mass = m1)
      sub_f2 <- matches %>% dplyr::select(feature_id = f2, adduct = a2, adduct_mass = m2)
      
      annotated_subtable <- bind_rows(sub_f1, sub_f2) %>%
        distinct() %>%
        left_join(sb %>% dplyr::select(feature_id, mzmed, rtmed), by = "feature_id") %>%
        mutate(neutral_mass = mzmed - adduct_mass) %>%
        dplyr::select(feature_id, mzmed, rtmed, adduct, neutral_mass)
      
      # --- THE FIX: CATCH THE LEFTOVERS ---
      # Find features in this group that didn't get an adduct match
      leftovers <- sb %>% filter(!feature_id %in% annotated_subtable$feature_id)
      
      if(nrow(leftovers) > 0) {
        leftover_subtable <- leftovers %>%
          dplyr::select(feature_id, mzmed, rtmed) %>%
          mutate(
            adduct = "unk",
            neutral_mass = mzmed + 1.007276 # Assuming [M-H]- for unknowns
          )
        # Bind the annotated pairs and the leftovers together
        subtable <- bind_rows(annotated_subtable, leftover_subtable) %>% arrange(mzmed)
      } else {
        subtable <- annotated_subtable %>% arrange(mzmed)
      }
      
    } else {
      # --- THE FIX: NO MATCHES FOUND AT ALL ---
      # If the group has >1 feature but no math works, treat ALL as singletons
      subtable <- sb %>%
        dplyr::select(feature_id, mzmed, rtmed) %>%
        mutate(
          adduct = "[M-H]-",
          neutral_mass = mzmed + 1.007276
        ) %>% arrange(mzmed)
    }
    
    # Save the complete group to the list
    group_rtmed <- median(subtable$rtmed)
    
    results_list[[as.character(gr)]] <- tibble(
      feature_group = gr,
      group_rtmed = group_rtmed,
      annotated_features = list(subtable),
      # This will now safely grab the M_mass of both real matches and leftovers
      M_mass = list(subtable$neutral_mass[which(subtable$adduct == "[M-H]-")]),
      n = nrow(sb)
    )
    
  } else {
    # ---------------------------------------------------------
    # TRUE SINGLETONS (Only 1 feature in the group)
    # ---------------------------------------------------------
    subtable <- tibble(
      feature_id = sb$feature_id, # Fixed typo here
      mzmed = sb$mzmed,
      rtmed = sb$rtmed, 
      adduct = "[M-H]-",
      neutral_mass = sb$mzmed + 1.007276
    )
    
    results_list[[as.character(gr)]] <- tibble(
      feature_group = gr,
      group_rtmed = subtable$rtmed,
      annotated_features = list(subtable),
      M_mass = list(subtable$neutral_mass[which(subtable$adduct == "[M-H]-")]),
      n = 1
    )
  }
}

# ---------------------------------------------------------
# 4. Create the Final Master Table
# ---------------------------------------------------------
final_table <- bind_rows(results_list)

# ---------------------------------------------------------
# 5. refine [M+H]+ identification (handle insource fragmentation)
# ---------------------------------------------------------
loss_deltas <- c(18.0106, 36.0212, 54.0318, 27.9949, 44.9976, 43.98983)
loss_labels <- c("[M-H2O-H]-", "[M-2H2O-H]-",  "[M-3H2O-H]-", "[M-CO-H]-", "[M-COOH-H]-", "[M-CO2-H]-" )
iso_deltas <- c(1.00335, 2.00671,3.01005)

cleaned_results <- list()

for(q in 1:nrow(final_table)) {
  
  row_data <- final_table[q, ]
  sub_df <- row_data$annotated_features[[1]]
  m_candidates <- sub_df%>%
    filter(adduct=="[M-H]-") %>%
    pull(mzmed) %>%
    sort()
  
  # If there are 0 or 1 [M+H]+ candidates, keep it as is and skip
  if(length(m_candidates) < 2) {
    cleaned_results[[q]] <- row_data
    next
  }
  
  # 1. ISOTOPES
  for (i in 1:(length(m_candidates)-1)) {
    for (j in (i+1):length(m_candidates)) {
      
      # Because we sorted, delta is always negative..inverte
      delta <- m_candidates[j] - m_candidates[i]
      
      if (any(abs(delta - iso_deltas) <= 0.005)) {
        true_M_iso <- m_candidates[i] 
        
        sub_df <- sub_df %>%
          mutate(
            adduct = case_when(
              # The extra "mzmed > true_M_iso" guarantees we only alter the heavier peak
              abs(mzmed - (true_M_iso + 1.00335)) <= 0.005 & mzmed > true_M_iso & adduct == "[M-H]-" ~ "[M+1-H]-",
              abs(mzmed - (true_M_iso + 2.00671)) <= 0.005 & mzmed > true_M_iso & adduct == "[M-H]-" ~ "[M+2-H]-",
              abs(mzmed - (true_M_iso + 3.01005)) <= 0.005 & mzmed > true_M_iso & adduct == "[M-H]-" ~ "[M+3-H]-",
              TRUE ~ adduct
            ),
            # Snap the neutral mass of the isotope back to the monoisotopic parent
            neutral_mass = ifelse(adduct %in% c("[M+1-H]-", "[M+2-H]-",  "[M+3-H]-"), true_M_iso, neutral_mass)
          )
      }
    }
  }
  
  # 2. DIMERS
  dimer_labels <- c("[2M-H]-", "[2M+Cl]-", "[2M+HCOO]-","[2M-2H+Na]-")
  dimer_additions <-c(-1.007276, 34.968853, 44.998203, 20.974666)
  
  for (i in 1:(length(m_candidates)-1)) {
    for (j in (i+1):length(m_candidates)) {
      
      # 1. Calculate the neutral mass of the lighter candidate (assumed monomer)
      true_M_monomer <- m_candidates[i] 
      true_neutral <- true_M_monomer + 1.007276 
      
      for (d in 1:length(dimer_labels)) {
        
        expected_dimer_mz <- (2 * true_neutral) + dimer_additions[d]
        
        if (abs(m_candidates[j] - expected_dimer_mz) <= 0.005) {
          
          target_dimer_label <- dimer_labels[d]
          sub_df <- sub_df %>%
            mutate(
              # Relabel the heavier peak to the correct dimer string
              adduct = ifelse(abs(mzmed - expected_dimer_mz) <= 0.005 & adduct == "[M-H]-", 
                              target_dimer_label, adduct),
              neutral_mass = ifelse(adduct == target_dimer_label, true_neutral, neutral_mass)
            )
        }
      } # End of dimer types loop
    }
  }
  
  # 3. INSOURCE FRAGMENTATION (Cascades)
  for (i in 1:(length(m_candidates)-1)) {
    for (j in (i+1):length(m_candidates)) {
      
      # Calculate the mass difference between the two candidates
      delta <- m_candidates[j] - m_candidates[i]
      
      # Loop through our 4 specific losses to see if there's a match
      for (k in 1:length(loss_deltas)) {
        
        if (abs(delta - loss_deltas[k]) <= 0.005) {
          fake_M_frag <- m_candidates[i]    # The lighter mass (fragment)
          true_M_cascade <- m_candidates[j] # The heavier mass (parent)
          target_label <- loss_labels[k]    # Pull the matching text label
          
          sub_df <- sub_df %>%
            mutate(
              # 1. Update the text label so we know EXACTLY what fell off
              adduct = ifelse(abs(mzmed - fake_M_frag) <= 0.005 & adduct == "[M-H]-", 
                              target_label, adduct),
              
              # 2. Snap the neutral mass up to the true parent 
              neutral_mass = ifelse(adduct == target_label, true_M_cascade, neutral_mass)
            )
        }
      } # End of the loss checking loop
      
    }
  }
  # 4. FINALIZE THE ROW
  
  # Update the subtable
  row_data$annotated_features[[1]] <- sub_df
  
  # Dynamically recalculate M_mass based on the features that SURVIVED as [M-H]-
  # This perfectly preserves any unrelated singletons that were co-eluting!
  row_data$M_mass[[1]] <- unique(sub_df$neutral_mass[sub_df$adduct == "[M+H]+"])
  
  cleaned_results[[q]] <- row_data
}

# Bind it all back together into a final, pristine table
cl_table_neg<- bind_rows(cleaned_results)

ftneg<- feat_def
data_neg <- fdata

save(cl_table_pos, cl_table_neg, file="rdata/alt_custom_pairing.RData")
```
plotting the adducts detected

```{r}
sel_pos <-cl_table_pos %>%
  dplyr::select(-M_mass,-n) %>% 
  unnest(cols = c(annotated_features)) %>%
    as.data.frame() %>%
  filter(adduct != "unk")%>%
    filter(adduct != "[M+H]+")
adduct_counts <- sel_pos %>%
  filter(adduct != "unk") %>%
  group_by(adduct) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(adduct = case_when(
    adduct %in% c("[M+1+H]+", "[M+2+H]+", "[M+3+H]+") ~ "Isotopes",
    TRUE ~ adduct
  )) %>%
  group_by(adduct) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  arrange(desc(n))

 p<- ggplot(adduct_counts, aes(x = reorder(adduct, n), y = n, fill = adduct)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(y = "Count", x = "Adduct", title="Positive ESI") +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal(base_size = 14)

sel_neg <-cl_table_neg %>%
  dplyr::select(-M_mass,-n) %>% 
  unnest(cols = c(annotated_features)) %>%
  as.data.frame()%>%
  filter(adduct != "unk") %>%
    filter(adduct != "[M-H]-")

adduct_counts <- sel_neg %>%
  filter(adduct != "unk") %>%
  group_by(adduct) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(adduct = case_when(
    adduct %in% c("[M+1-H]-", "[M+2-H]-", "[M+3-H]-") ~ "Isotopes",
    TRUE ~ adduct
  )) %>%
  group_by(adduct) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  arrange(desc(n))

 n<- ggplot(adduct_counts, aes(x = reorder(adduct, n), y = n, fill = adduct)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(y = "Count", x = "Adduct", title="Negative ESI") +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal(base_size = 14)
 
plot_grid(p, n, ncol = 2)
```

#NEUTRAL

making matching between all positive and negative feature, considering only M+H and M-H pairs

```{r}
#extracting feature with protonated or deprotonated forms (excluding isotopes and adducts)
sel_pos <-cl_table_pos %>%
  dplyr::select(-M_mass,-n) %>% 
  unnest(cols = c(annotated_features)) %>%
    as.data.frame() %>%
  filter(adduct %in% c("[M+H]+"))

sel_neg <-cl_table_neg %>%
  dplyr::select(-M_mass,-n) %>% 
  unnest(cols = c(annotated_features)) %>%
  as.data.frame()%>%
  filter(adduct %in% c("[M-H]-"))

ft_match <- matchValues(
  sel_pos, sel_neg, param = MzRtParam(ppm = 10, toleranceRt = 5),
  mzColname = c("neutral_mass", "neutral_mass"),
  rtColname  = c("rtmed", "rtmed")  #prima query e dopo tabella target
  )

tb_match <- matchedData(ft_match) %>%
  as.data.frame()
tb_match <- tb_match[!is.na(tb_match$target_mzmed),]

#handling double matching
#ensuring the difference of original m/z is 2.01455

tb_match<- tb_match %>%
  group_by(feature_id) %>%
  filter(
    # Keep the row IF it is the only match (not duplicated)
    n() == 1 | 
    # OR keep the row IF it passes your 2.01455 Da distance check
    abs(abs(target_mzmed - mzmed) - 2.01455) <= 0.005
  ) %>%
  ungroup() %>%
  as.data.frame() %>%
  dplyr::select(feature_group, feature_id, group_rtmed, mzmed, adduct, target_feature_id, target_group_rtmed, target_mzmed, target_adduct )

```

refining the matches according to inter-polarity correlation of intensities

```{r}
#filtering for matches which has a correlation between negative and positive
vals_neg <- featureValues(data_neg) %>%
    as.data.frame()
colnames(vals_neg) <- paste0(sampleData(data_neg)$sample,"_", sampleData(data_neg)$rep)
vals_neg<- vals_neg[,grepl("A", colnames(vals_neg))]
vals_pos <- featureValues(data_pos) %>%
    as.data.frame()
colnames(vals_pos) <- paste0(sampleData(data_pos)$sample,"_", sampleData(data_pos)$rep)
vals_pos<- vals_pos[,grepl("A", colnames(vals_pos))]

#matching sample order in different polarity
common_samples <- intersect(colnames(vals_neg), colnames(vals_pos))
vals_neg <- vals_neg[, common_samples]
vals_pos <- vals_pos[, common_samples]

tb_match$correlation <- NA

for(i in 1:nrow(tb_match)){
  neg_id <- tb_match$target_feature_id[i]
  pos_id <- tb_match$feature_id[i]
  
  vln <- vals_neg[neg_id, ] %>%
    log2()
  vlp <- vals_pos[pos_id,] %>%
    log2()
  
  valid_indices <- !is.na(vlp) & !is.na(vln)
  vln <- vln[valid_indices]
  vlp <- vlp[valid_indices]
  
  cor_score<- cor(vln, vlp, method = "pearson")
  tb_match$correlation[i] <- cor_score
}

hist(tb_match$correlation)

tb_match <- tb_match %>%
  filter(abs(correlation) > 0.7)

paste("match with cromatographic correlation:", nrow(tb_match))

#in case of multiple matching select the one with highest correlation across samples
#for positive ionization
dup<- unique(tb_match$feature_id[which(duplicated(tb_match$feature_id))])

if(length(dup)>0){
 
tb_match<- tb_match %>%
  group_by(feature_id) %>%
  filter(
    # Keep the row IF it is the only match (not duplicated)
    n() == 1 | 
    # OR keep the row IF it passes your 2.01455 Da distance check
   row_number() == which.max(abs(correlation))
  ) %>%
  ungroup()

}
#for negative ionization
dup<- unique(tb_match$target_feature_id[which(duplicated(tb_match$target_feature_id))])

if(length(dup)>0){
 
tb_match<- tb_match %>%
  group_by(target_feature_id) %>%
  filter(
    # Keep the row IF it is the only match (not duplicated)
    n() == 1 | 
    # OR keep the row IF it passes your 2.01455 Da distance check
   row_number() ==  which.max(abs(correlation))
  ) %>%
  ungroup()

}

paste("match without double results:", nrow(tb_match))

tb_match$n_mass <- (tb_match$mzmed+tb_match$target_mzmed)/2
comonft <- tb_match %>%
  dplyr::select(group_rtmed, n_mass) %>%
  rename("group_rtmed"="rtmed")%>%
  as.data.frame()

comonft$name_p <- tb_match$feature_id
comonft$name_n <- tb_match$target_feature_id
nrow(comonft)
```

selecting and cleaning MSMS spectra

```{r warning=FALSE}
register(SerialParam())
MS2pos <- spspos[which(spspos$feature_id %in% comonft$name_p)] %>%
  cleaning_ref()
MS2neg <- spsneg[which(spsneg$feature_id %in% comonft$name_n)] %>%
  cleaning_ref()

#precurosr removal

MS2neg <- filterPrecursorPeaks(MS2neg, ppm = 50, mz = ">=")
MS2pos <- filterPrecursorPeaks(MS2pos, ppm = 50, mz = ">=")

length(MS2pos)
length(MS2neg)
```

keep MS 2 relative to features detected in MS1 and making a single spectra for negative and positive

```{r warning=FALSE}
#function to handle overlapping ions
process_data <- function(data) {
  data <- data %>%
    arrange(mz) %>%
    mutate(group = cumsum(c(TRUE, diff(mz) >= 0.001))) %>%
    group_by(group) %>%
    summarise(
      mz = mean(mz),
      intensity = sum(intensity)
    ) %>%
    ungroup() %>%
    dplyr::select(-group)
  return(data)
} 

#NEUTRAL creation
    
    #preparing metadata
    meta <- comonft %>%
      dplyr::select(rtmed, name_p, name_n) 
    meta$feature_id <- meta$name_p
    
    meta$precursorMz <- NA
    meta$mz <- meta$intensity <- vector(mode="list", length=nrow(meta))
    
    for(i in 1:nrow(meta)){
      register(SerialParam())
      
      #positive
      akap <- MS2pos[which(MS2pos$feature_id == meta$feature_id[i])]
      if(length(akap)>0){
        mzpos <- mz(akap)[[1]] 
        mzpos <- mzpos - 1.00728
        intpos <- intensity(akap)[[1]]
        intpos <- (intpos/max(intpos))*100
        df_pos <- data.frame(mz = mzpos, intensity = intpos)
        
        #negative
        akan <- MS2neg[which(MS2neg$feature_idid == meta$name_n[i])]
        if(length(akan)>0){
          mzneg <- mz(akan)[[1]] 
          mzneg <- mzneg + 1.00728
          intneg <- intensity(akan)[[1]]
          intneg <- (intneg/max(intneg))*100
          df_neg <- data.frame(mz = mzneg, intensity = intneg) 
          
          #both
          df <- bind_rows(df_neg, df_pos) %>%
            arrange(mz)
          df<- as.data.frame(process_data(df))
          meta$mz[[i]]<- as.numeric(df$mz)
          meta$intensity[[i]] <- as.numeric(df$intensity)
          
          meta$precursorMz[i] <- mean(c(precursorMz(akan),precursorMz(akap)))
        }
      }
      meta_all <- rbind(meta, meta_all)
    }

MS2neu <- Spectra(meta_all)

save(comonft, tb_match, spsneg, spspos,MS2pos, MS2neg, MS2neu, file="rdata/alt_msms2.RData")
```

calculating similarity with GNPS similarity function

```{r}
matrix <- compareSpectra(MS2neu, MAPFUN = joinPeaksGnps, tolerance = 0.01, FUN = MsCoreUtils::gnps)
colnames(matrix)<- MS2neu$feature_id
rownames(matrix)<- MS2neu$feature_id
matrix_pos <- compareSpectra(MS2pos, MAPFUN = joinPeaksGnps, tolerance = 0.01, FUN = MsCoreUtils::gnps)
colnames(matrix_pos)<- MS2pos$feature_id
rownames(matrix_pos)<- MS2pos$feature_id
matrix_neg <- compareSpectra(MS2neg, MAPFUN = joinPeaksGnps, tolerance = 0.01, FUN = MsCoreUtils::gnps)
colnames(matrix_neg)<- MS2neg$feature_id
rownames(matrix_neg)<- MS2neg$feature_id

save(comonft, matrix, matrix_pos, matrix_neg, tb_match, spsneg, spspos,MS2pos, MS2neg, MS2neu, file="rdata/alt_msms.RData")
```

#MNs

names to highlighte

```{r}
#LT toxin names (3 isomers)

tname_p <- comonft$name_p[which(MsCoreUtils::between(comonft$n_mass, 282.1466+0.001*c(-1,1))&
                   MsCoreUtils::between(comonft$rtmed, c(240,280)))]

tname_n <- comonft$name_n[which(comonft$name_p %in% tname_p)]
```

neutral molecualr network (interactive)

```{r}
threshold <- 0.7

adj_matrix <- matrix
adj_matrix[adj_matrix < threshold] <- 0
diag(adj_matrix) <- 0

g <- graph_from_adjacency_matrix(
  adj_matrix,
  mode = "undirected",
  weighted = TRUE,
  diag = FALSE
)

nodes <- data.frame(
  id    = V(g)$name)

mtc <- match(nodes$id,
             comonft$name_p)

nodes$rt <- round(comonft$rtmed[mtc]/60,2)
nodes$mz<- round(comonft$n_mass[mtc],4)
nodes$title <- paste(nodes$id,"\n",nodes$mz,"@",nodes$rt)

nodes$color <- ifelse(nodes$id %in% tname_p, "indianred", "lightgray")

edges <- data.frame(
  from = ends(g, E(g))[,1],
  to   = ends(g, E(g))[,2]
)

mn <- visNetwork(nodes, edges) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = list(enabled = TRUE, degree = 1)) %>%
  visNodes(shape = "dot", borderWidth = 0) %>%
  visEdges(selectionWidth = 3) %>%
  visPhysics(solver = "forceAtlas2Based", 
             forceAtlas2Based = list(springLength = 100))

mn
```
