---
title: "merging_poneg"
output: html_document
date: "2024-06-27"
---

```{r}
library(xcms)
library(effectsize)
library(dplyr)
library(tidyr)
library(purrr)
library(rlang)
library(MsCoreUtils)
library(Spectra)
library(MsBackendMgf)
library(readr)
library(stringr)
library(MetaboCoreUtils)
library(igraph)
library(visNetwork)
library(MsExperiment)
library(RuSirius)
library(stats)
library(MetaboAnnotation)
library(RColorBrewer)
library(MASS)

cleaning_ref<- function(ms2){
  ms2_tb <- as.data.frame(
    spectraData(ms2) 
  ) %>%
    mutate( ms2_raw = mapply(cbind, mz(ms2), intensity(ms2), SIMPLIFY = FALSE)) %>%
    mutate(ms2_raw = map(ms2_raw, function(t){
      colnames(t) <- c("mz", "intensity")
      t %>%
        as.data.frame() %>%
        mutate(mz = as.numeric(mz)) %>%
        mutate(intensity = as.numeric(intensity))
    })) %>%
    mutate(ms2_RLM = map(ms2_raw, function(t){
      t <- t %>%
        filter(intensity>0) %>%
        arrange(intensity) %>%
        mutate(order = seq(nrow(.)))
      p75 <- t %>%
        pull(intensity) %>% 
        quantile(0.75)
      u <- t %>%
        filter(intensity < p75)
      if(nrow(u)>3){
        mdl <- rlm(data = u, log10(intensity) ~ order)
        t_filt<- t %>% 
          mutate(pred = predict(mdl, t)) %>%
          mutate(res = abs(pred - log10(intensity))) %>%
          filter(res > sd(residuals(mdl))*3) %>%
          arrange(mz)
        if (nrow(t_filt) > 0) {
          t<- t_filt
        } else {
          t<- t %>% arrange(mz)
        }
      } else {
        return(t %>% arrange(mz))  # Not enough points to model
      }
    }))
  ms2_cl <- ms2_tb %>%
    mutate(
      mz = map(ms2_RLM, ~ .x$mz),
      intensity = map(ms2_RLM, ~ .x$intensity)
    ) %>%
    filter(map_lgl(mz, ~ length(.x) > 0)) %>%
    as.data.frame() %>%
    Spectra()
  return(ms2_cl)
}
```

#POSITIVE XCMS

loading files 
using only DDA data, also for peak peaking, with metadata

```{r}
pth <- file.path( "alternaria_data/pos/data/mzML")
fls <- list.files(pth, pattern = "mzML", full.names = TRUE)
fls <- fls[!grepl("blank|MED|MEDW", fls, ignore.case = TRUE)]
fls <- fls[grep("DDA", fls)]
```

peak piking 
 
```{r}
mse <- readMsExperiment(spectraFiles = fls)

#filter wrong spectra
idx <- which(round(spectra(mse)$precursorMz) != round(spectra(mse)$isolationWindowTargetMz) & spectra(mse)$msLevel==2)
filterByIndex <- function(x, index) { x[index] }
filterSpectra(mse, filterByIndex, -idx)

cwp<- CentWaveParam(ppm = 10,
                    peakwidth = c(3, 10),
                     prefilter = c(5, 1000000),
                     noise = 50000)
data <- findChromPeaks(mse, param = cwp) 
```
 
 features creation
 
```{r}
#' Peak refinement
mnp <- MergeNeighboringPeaksParam(expandRt = 2, expandMz = 0.001,
                                  minProp = 3/4)
data <- refineChromPeaks(data, param = mnp)
#' Alignment
sampleData(data)$highRes <- TRUE
pdp1 <- PeakDensityParam(sampleGroups = sampleData(data)$highRes, bw = 2,
                         minFraction = 0.8, binSize = 0.01)
data <- groupChromPeaks(data, param = pdp1)
pgp <- PeakGroupsParam(minFraction = 0.8, extraPeaks = 100, span = 0.8)
data <- adjustRtime(data, param = pgp)
#' Correspondence analysis
pdp2 <- PeakDensityParam(sampleGroups =sampleData(data)$highRes, bw = 2,
                         minFraction = 0.3, binSize = 0.01) 
data <- groupChromPeaks(data, param = pdp2)
#' Gap-filling
data <- fillChromPeaks(data, param = ChromPeakAreaParam())
ftvalues <-data.frame(featureValues(data, value = "into", method="sum"))
ftdetec <- data.frame(featureDefinitions(data))
ftvalues <- ftvalues[rownames(ftvalues) %in% rownames(ftdetec),]
ftdetec$meanval <- colMeans(t(ftvalues), na.rm = TRUE)

```
 
searching for msms related to each features
 
```{r}
ms2_spectra <- featureSpectra(data, expandRt = 0, return.type = "Spectra",
                              method ="largest_bpi",
                              ppm = 5, 
                              featureColumns = c("mzmed", "rtmed", "rtmin", "rtmax"))
ms2_allspectra_pos <- featureSpectra(data, expandRt = 0, return.type = "Spectra",
                              ppm = 5, 
                              featureColumns = c("mzmed", "rtmed", "rtmin", "rtmax"))

ms2_spectra$id <- paste0(basename(ms2_spectra$dataOrigin), ms2_spectra$scanIndex)

#features with at leat one fragmentation spectra
features <- unique(spectraData(ms2_spectra)$feature_id)

```

selecting one ms2 for each feature, selecting the one with highest base peak intensity

```{r}

sp <- split(spectraData(ms2_spectra), spectraData(ms2_spectra)$feature_id)

cp <- vapply(sp, function(x){
    x$chrom_peak_id[which.max(x$basePeakIntensity)]
}, character(1))
ms2_spectra <- ms2_spectra[which(spectraData(ms2_spectra)$chrom_peak_id %in% cp)]

#Noise cleaning
ms2_spectra <- cleaning_ref(ms2_spectra)

spspos<- ms2_spectra
ftdetec$name <- rownames(ftdetec)
ftpos <- ftdetec[which(ftdetec$name %in% spspos$feature_id),]
length(spspos)
nrow(ftpos)
```

#NEGATIVE XCMS

loading files 
using only DDA data, also for peak peaking

```{r}
pth <- file.path( "alternaria_data/neg/data/mzML")
fls <- list.files(pth, pattern = "mzML", full.names = TRUE)
fls <- fls[!grepl("blank|MED|MEDW", fls, ignore.case = TRUE)]
fls <- fls[grep("DDA", fls)]
```

peak piking 
 
```{r}
mse <- readMsExperiment(spectraFiles = fls)

#filter wrong spectra
idx <- which(round(spectra(mse)$precursorMz) != round(spectra(mse)$isolationWindowTargetMz) & spectra(mse)$msLevel==2)
filterByIndex <- function(x, index) { x[index] }
filterSpectra(mse, filterByIndex, -idx)

cwp<- CentWaveParam(ppm = 10,
                    peakwidth = c(3, 10),
                     prefilter = c(5, 500000),
                     noise = 50000)
data <- findChromPeaks(mse, param = cwp) 
```
 
 features creation
 
```{r}
#' Peak refinement
mnp <- MergeNeighboringPeaksParam(expandRt = 2, expandMz = 0.001,
                                  minProp = 3/4)
data <- refineChromPeaks(data, param = mnp)
#' Alignment
sampleData(data)$highRes <- TRUE
pdp1 <- PeakDensityParam(sampleGroups = sampleData(data)$highRes, bw = 2,
                         minFraction = 0.8, binSize = 0.01)
data <- groupChromPeaks(data, param = pdp1)
pgp <- PeakGroupsParam(minFraction = 0.8, extraPeaks = 100, span = 0.8)
data <- adjustRtime(data, param = pgp)
#' Correspondence analysis
pdp2 <- PeakDensityParam(sampleGroups =sampleData(data)$highRes, bw = 2,
                         minFraction = 0.3, binSize = 0.01) 
data <- groupChromPeaks(data, param = pdp2)
#' Gap-filling
data <- fillChromPeaks(data, param = ChromPeakAreaParam())
ftvalues <-data.frame(featureValues(data, value = "into", method="sum"))
ftdetec <- data.frame(featureDefinitions(data))
ftvalues <- ftvalues[rownames(ftvalues) %in% rownames(ftdetec),]
ftdetec$meanval <- colMeans(t(ftvalues), na.rm = TRUE)
```


searching for msms realted to each features
 
```{r}
ms2_spectra <- featureSpectra(data, expandRt = 0, return.type = "Spectra",
                              method ="largest_bpi",
                              ppm = 5, 
                              featureColumns = c("mzmed", "rtmed", "rtmin", "rtmax"))

ms2_allspectra_neg <- featureSpectra(data, expandRt = 0, return.type = "Spectra",
                              ppm = 5, 
                              featureColumns = c("mzmed", "rtmed", "rtmin", "rtmax"))


ms2_spectra$id <- paste0(basename(ms2_spectra$dataOrigin), ms2_spectra$scanIndex)

#features with at leat one fragmentation spectra
features <- unique(spectraData(ms2_spectra)$feature_id)

```
 
 selecting one msms for each features
 
```{r}
sp <- split(spectraData(ms2_spectra), spectraData(ms2_spectra)$feature_id)

cp <- vapply(sp, function(x){
    x$chrom_peak_id[which.max(x$basePeakIntensity)]
}, character(1))
ms2_spectra <- ms2_spectra[which(spectraData(ms2_spectra)$chrom_peak_id %in% cp)]

#Noise cleaning
ms2_spectra <- cleaning_ref(ms2_spectra)

spsneg<- ms2_spectra
ftdetec$name <- rownames(ftdetec)
ftneg <- ftdetec[which(ftdetec$name %in% spsneg$feature_id),]
length(spsneg)
nrow(ftneg)

```

#NEUTRAL

```{r}
ftpos <- ftpos %>%
  filter(rtmed>30&rtmed<600)
ftpos<- ftpos%>%
  filter(log10(meanval)>6)

ftneg <- ftneg %>%
  filter(rtmed>30&rtmed<600)
ftneg <- ftneg %>%
  filter(log10(meanval)>6)

#create a table with correct mass
ftpos$neutralmass <- ftpos$mzmed - 1.00728
ftneg$neutralmass <- ftneg$mzmed + 1.00728
```

selecting only common ft between positive and negative ionization, match for rt and mz 

```{r}

ftpos <- as.data.frame(ftpos)
ftneg <- as.data.frame(ftneg)
ft_match <- matchValues(
  ftneg, ftpos, param = MzRtParam(ppm = 5, toleranceRt = 5),
  mzColname = c("neutralmass", "neutralmass"),
  rtColname  = c("rtmed", "rtmed")  
  )
tb_match <- matchedData(ft_match)
tb_match <- tb_match[!is.na(tb_match$target_mzmed),]
tb_match <- tb_match[!grepl("\\.", rownames(tb_match)), ]  #delete double matches
tb_match <- as.data.frame(tb_match)
tb_match <- tb_match %>%
  distinct(target_name, .keep_all = TRUE)
tb_match$n_mass <- (tb_match$neutralmass+tb_match$target_neutralmass)/2
comonft <- tb_match %>%
  dplyr::select(all_of(c("rtmed", "n_mass", "meanval")))
comonft$name_p <- tb_match$target_name
comonft$name_n <- tb_match$name
```

selecting MS2 of fature used for NMN

```{r warning=FALSE}
MS2pos <- spspos[which(spspos$feature_id %in% comonft$name_p)]
MS2neg <- spsneg[which(spsneg$feature_id %in% comonft$name_n)]

length(MS2neg)
length(MS2pos)
```

single spectra for negative and positive

```{r warning=FALSE}
process_data <- function(data) {
  data <- data %>%
    arrange(mz) %>%
    mutate(group = cumsum(c(TRUE, diff(mz) >= 0.0005))) %>%
    group_by(group) %>%
    summarise(
      mz = mean(mz),
      intensity = mean(intensity)
    ) %>%
    ungroup() %>%
    dplyr::select(-group)
  return(data)
} 

comonft$spectrum <- NA
neutral_spectra <- Spectra()


for(i in 1:nrow(comonft)){
  register(SerialParam())
  #positive
  fl <- MS2pos
  akap <- fl[i]
    if(length(akap)>0){
    mzpos <- mz(akap)[[1]] 
    mzpos <- mzpos - 1.00728
    intpos <- intensity(akap)[[1]]
    intpos <- (intpos/max(intpos))*100
    df_pos <- data.frame(mz = mzpos, intensity = intpos)
    #negative
    fl <- MS2neg
    neg_name <- comonft$name_n[which(comonft$name_p==akap$feature_id)]
    akan <- fl[which(fl$feature_id==neg_name)]
        if(length(akan)>0){
      mzneg <- mz(akan)[[1]] 
      mzneg <- mzneg + 1.00728
      intneg <- intensity(akan)[[1]]
      intneg <- (intneg/max(intneg))*100
      df_neg <- data.frame(mz = mzneg, intensity = intneg) 
          
      #both
      df <- bind_rows(df_neg, df_pos) %>%
        arrange(mz)
      df<- as.data.frame(process_data(df))
      totmz<- as.numeric(df$mz)
      totint <- as.numeric(df$intensity)
      
      prec<- mean(c(precursorMz(akan),precursorMz(akap)))
      
      spectrum<- data.frame(
        id= akap$feature_id,
        mz = I(list(totmz)),
        intensity = I(list(totint)),
        precursorMz=prec
      )
      sps <- Spectra(spectrum)
      neutral_spectra <- c(neutral_spectra, sps)
    }
  }
  print(i)
}

```

#modualrity mesurements 

global and local modularity for the toxin-related isomers across different similarity threshold (from 0 to 1)
The mesurements where repeated multiple times (25) in order to consider variability. For each repetition, a MS/MS spectra for each feature where randomly chose within feature related spectra. 
first we identify our subset of features that are representing the same compound (isomers)

```{r}
#manual identification 
toxp <- "FT1776"
toxn <- "FT1104"
isop <- c("FT1775","FT1777")
ison <- c("FT1103","FT1105")
#identification of dehydratation of toxin isomers
sub<- comonft[which(MsCoreUtils::between(comonft$rtmed, c(4*60,4.5*60))),]
sub<- sub[which(MsCoreUtils::between(sub$n_mass, 264.1360+0.0005*c(-1,1))),]
allisop <- c(isop, sub$name_p, toxp)
allison <- c(ison, sub$name_n, toxn)
```

define the function for local modularity calculation

```{r}
LQ <- function(graph, allisop){
  sel_nodes <- V(graph)[name %in% allisop]
#identify my cluster (toxin isomers)
sub <- induced_subgraph(graph, vids = allisop)
#edges within my cluster
Li<- sum(E(sub)$weight)
#edges with neighbour and within my cluster
vicini<- unique(unlist(neighborhood(graph, order = 1, nodes = allisop)))
local_subgraph <- induced_subgraph(graph, vids = vicini)
L_N <- sum(E(local_subgraph)$weight)
#edges external to my cluster
df1<- as_data_frame(local_subgraph, what="edges") %>%
  filter(from %in% allisop |to %in% allisop)
df2<- as_data_frame(local_subgraph, what="edges") %>%
  filter(from %in% allisop &to %in% allisop)
df <- anti_join(df1, df2, by = c("from", "to"))

L_out <- sum(df$weight)

LQ <- (Li / L_N) - ((Li * L_out) / (L_N^2))

return(LQ)
}
```

calualtion of mosularities across similarity thresholds

```{r}
ms2_spectra_neg<- ms2_allspectra_neg    #all negative MS/MS
ms2_spectra_pos <- ms2_allspectra_pos    #all positive MS/MS
#giving an id for changing spectra within a feature
ms2_spectra_neg$idf <- paste0(ms2_spectra_neg$scanIndex, basename(ms2_spectra_neg$dataOrigin))
ms2_spectra_neg<- ms2_spectra_neg[!duplicated(ms2_spectra_neg$idf)]
ms2_spectra_pos$idf <- paste0(ms2_spectra_pos$scanIndex, basename(ms2_spectra_pos$dataOrigin))
ms2_spectra_pos<- ms2_spectra_pos[!duplicated(ms2_spectra_pos$idf)]

metadata_neg <- as.data.frame(spectraData(ms2_spectra_neg))
metadata_pos <- as.data.frame(spectraData(ms2_spectra_pos))

valtab <- data.frame(rep=1:25)
valtab$data <- vector(mode="list", length=nrow(valtab))
valtab$LQ <- vector(mode="list", length=nrow(valtab))

#loop for differnet validation

for(w in 1:nrow(valtab)){
#selecting neg sps
  tmpneg<- Spectra()
unique_features <- metadata_neg %>%
  group_by(feature_id) %>%
  slice_sample(n = 1) %>%
  ungroup() %>%
  filter(feature_id %in% MS2neg$feature_id)
tmpneg <- ms2_spectra_neg[match(unique_features$idf, ms2_spectra_neg$idf)]
tmpneg$id<- tmpneg$feature_id
#selecting pos sps
tmppos<- Spectra()
unique_features <- metadata_pos %>%
  group_by(feature_id) %>%
  slice_sample(n = 1) %>%
  ungroup() %>%
  filter(feature_id %in% MS2pos$feature_id)
tmppos <- ms2_spectra_pos[match(unique_features$idf, ms2_spectra_pos$idf)]
tmppos$id <- tmppos$feature_id
#making neutral
tmpneu <- Spectra()
for(i in 1:nrow(comonft)){
  register(SerialParam())
  #positive
  fl <- tmppos
  akap <- fl[which(fl$feature_id==comonft$name_p[i])]
  if(length(akap)>0){
    mzpos <- mz(akap)[[1]] - 1.00728
    intpos <- intensity(akap)[[1]]
    intpos <- (intpos/max(intpos))*100
    df_pos <- data.frame(mz = mzpos, intensity = intpos)
    #negative
    fl <- tmpneg
    akan <- fl[which(fl$feature_id==comonft$name_n[i])]
    if(length(akan)>0){
      mzneg <- mz(akan)[[1]] + 1.00728
      intneg <- intensity(akan)[[1]]
      intneg <- (intneg/max(intneg))*100
      df_neg <- data.frame(mz = mzneg, intensity = intneg) 
      
      #both
      df <- bind_rows(df_neg, df_pos) %>%
        arrange(mz)
      df<- as.data.frame(process_data(df))
      totmz<- as.numeric(df$mz)
      totint <- as.numeric(df$intensity)
      
      prec<- max(c(precursorMz(akan),precursorMz(akap)))
      precint<- mean(precursorIntensity(akap)+precursorIntensity(akan))
      
      spectrum<- data.frame(
        id= comonft$name_p[i],
        mz = I(list(totmz)),
        intensity = I(list(totint)),
        precursorMz=prec,
        precursorIntensity=precint,
        rtime <- comonft$rtmed[i]
      )
      spectrum <- Spectra(spectrum)
      tmpneu <- c(tmpneu, spectrum)
    }
  }
}
register(MulticoreParam(detectCores() - 1))  
#create matrices
mat_neu <- compareSpectra(tmpneu, MAPFUN = joinPeaksGnps, tolerance = 0.01, FUN = MsCoreUtils::gnps, BPPARAM = MulticoreParam())
mat_neg <- compareSpectra(tmpneg, MAPFUN = joinPeaksGnps, tolerance = 0.01, FUN = MsCoreUtils::gnps, BPPARAM = MulticoreParam())
mat_pos <- compareSpectra(tmppos, MAPFUN = joinPeaksGnps, tolerance = 0.01, FUN = MsCoreUtils::gnps, BPPARAM = MulticoreParam())

seq <- seq(0,1,0.05)  #simialrity threshold for the mesurements

#df for global modualrity
mod_thr <- data.frame(n=seq,   
                      neu=NA,
                      pos=NA,
                      neg=NA)
#df for local modularity 
LQ_thr <- data.frame(n=seq,
                     neu=NA,
                     pos=NA,
                     neg=NA)

#nodes to consider
all_nodes <- data.frame(
  id = neutral_spectra$id
)
all_nodes_p <- data.frame(
  id = MS2pos$feature_id
)
all_nodes_n <- data.frame(
  id = MS2neg$feature_id
)

#making MNs an calculate modularity and LQ for toxin isomers

for(i in 1:length(seq)){
  thr <- seq[i]
  #neu
  edges<- create_edge_list(mat_neu, thr, tmpneu)
  gneu <- graph_from_data_frame(d = edges, vertices = all_nodes, directed = FALSE)
  V(gneu)$iso<- NA
  V(gneu)$iso[which(V(gneu)$name%in% allisop)]<- "toxin"
  V(gneu)$community <- ifelse(V(gneu)$name %in% allisop, 1, 2)
  mod_thr$neu[i]<- modularity(gneu, membership = V(gneu)$community)
  LQ_thr$neu[i]<- LQ(gneu, allisop)
  #pos
  edges<- create_edge_list(mat_pos, thr, tmppos)
  gpos <- graph_from_data_frame(d = edges, vertices = all_nodes_p, directed = FALSE)
  V(gpos)$iso<- NA
  V(gpos)$iso[which(V(gpos)$name%in%allisop)]<- "toxin"
  V(gpos)$community <- ifelse(V(gpos)$name %in% allisop, 1, 2)
  mod_thr$pos[i]<-modularity(gpos, membership = V(gpos)$community)
  LQ_thr$pos[i]<- LQ(gpos, allisop)
  #neg
  edges<- create_edge_list(mat_neg, thr, tmpneg)
  gneg <- graph_from_data_frame(d = edges, vertices = all_nodes_n, directed = FALSE)
  V(gneg)$iso<- NA
  V(gneg)$iso[which(V(gneg)$name%in%allison)]<- "toxin"
  V(gneg)$community <- ifelse(V(gneg)$name %in% allison, 1, 2)
  mod_thr$neg[i]<- modularity(gneg, membership = V(gneg)$community)
  LQ_thr$neg[i]<- LQ(gneg, allison)
}
valtab$data[[w]] <-mod_thr
valtab$LQ[[w]] <-LQ_thr
print(w)
}

```




