---
title: "library_workflow"
author: "Nicola Dalla Valle"
date: "2025-08-25"
output: html_document
---

code used for libraies file extraction with the following MS/MS selection and neutral pseudo-spectra creation for compound with both polarities data.

```{r}
library(dplyr)
library(tidyverse)
library(purrr)
library(stringr)
library(jsonlite)
library(Spectra)
library(MSnbase)
library(BiocParallel)
library(sampling)
library(MASS)

process_data <- function(data) {
  data <- data %>%
    arrange(mz) %>%
    mutate(group = cumsum(c(TRUE, diff(mz) >= 0.0005))) %>%
    group_by(group) %>%
    summarise(
      mz = mean(mz),
      intensity = mean(intensity)
    ) %>%
    ungroup() %>%
    dplyr::select(-group)
  return(data)
} 

create_edge_list <- function(similarity_matrix, thr, spectra) {
  non_zero_indices <- which(similarity_matrix > thr, arr.ind = TRUE)
  if (length(non_zero_indices) > 0 && is.null(dim(non_zero_indices))) {
    non_zero_indices <- matrix(non_zero_indices, ncol = 2, byrow = TRUE)
  }
  if (length(non_zero_indices) == 0) {
    return(data.frame(from = character(), to = character(), weight = numeric()))
  }
  non_zero_indices <- non_zero_indices[non_zero_indices[, 1] < non_zero_indices[, 2], , drop = FALSE]
  if (nrow(non_zero_indices) == 0) {
    return(data.frame(from = character(), to = character(), weight = numeric()))
  }
  # Create an edge list from these indices
  edges <- data.frame(
    from = spectra$id[non_zero_indices[, 1]],
    to = spectra$id[non_zero_indices[, 2]],
    weight = similarity_matrix[non_zero_indices]
  )
  
  return(edges)
}

cleaning_ref<- function(ms2){
   if (length(mz(ms2)) <= 30) {  #apply cleaning for MS/MS with more than 30 ions
    return(ms2)
  }
ms2_tb <- as.data.frame(
    spectraData(sps) 
  ) %>%
    mutate( ms2_raw = mapply(cbind, mz(ms2), intensity(ms2), SIMPLIFY = FALSE)) %>%
    mutate(ms2_raw = map(ms2_raw, function(t){
      colnames(t) <- c("mz", "intensity")
      t %>%
        as.data.frame() %>%
        mutate(mz = as.numeric(mz)) %>%
        mutate(intensity = as.numeric(intensity))
    })) %>%
    mutate(ms2_RLM = map(ms2_raw, function(t){
      t <- t %>%
        filter(intensity>0) %>%
        arrange(intensity) %>%
        mutate(order = seq(nrow(.)))
      p75 <- t %>%
        pull(intensity) %>% 
        quantile(0.75)
      u <- t %>%
        filter(intensity < p75)
      if(nrow(u)>3){
        mdl <- rlm(data = u, log10(intensity) ~ order)
        t_filt<- t %>% 
          mutate(pred = predict(mdl, t)) %>%
          mutate(res = abs(pred - log10(intensity))) %>%
          filter(res > sd(residuals(mdl))*3) %>%
          arrange(mz)
        if (nrow(t_filt) > 0) {
          t<- t_filt
        } else {
          t<- t %>% arrange(mz)
        }
      } else {
        return(t %>% arrange(mz))  # Not enough points to model
      }
    }))
  ms2_cl <- ms2_tb %>%
    mutate(
      mz = map(ms2_RLM, ~ .x$mz),
      intensity = map(ms2_RLM, ~ .x$intensity)
    ) %>%
    filter(map_lgl(mz, ~ length(.x) > 0)) %>%
    as.data.frame() %>%
    Spectra()
  return(ms2_cl)
}


```

json library from MONA repository

```{r}
#assuming our libraries is downloaded ad .JSON file in "path" variable
fls <- fromJSON(path)
fls <- fls %>%
  dplyr::select("compound","id", "metaData", "annotations", "spectrum")
```

extracting interesting variables from dataset

```{r}
fls$pol <- sapply(fls$metaData, function(x) {
  x<- as.data.frame(x)    
  ionization_row <- which(x$name == "ionization mode")
  return(x$value[ionization_row])
       })
fls$adduct <- sapply(fls$metaData, function(x) {
  x<- as.data.frame(x)    
  ionization_row <- which(x$name == "precursor type")
  return(x$value[ionization_row])
       })
fls$precursorMz <- sapply(fls$metaData, function(x) {
  x<- as.data.frame(x)    
  ionization_row <- which(x$name == "precursor m/z")
  return(x$value[ionization_row])
       })
fls$instrument <- sapply(fls$metaData, function(x) {
  x<- as.data.frame(x)    
  ionization_row <- which(x$name == "instrument type")
  return(x$value[ionization_row])
       })
fls$entropy <- sapply(fls$metaData, function(x) {
  x<- as.data.frame(x)    
  ionization_row <- which(x$name == "spectral entropy")
  return(x$value[ionization_row])
       })
fls$inchi <- sapply(fls$compound, function(x){
  tab <- as.data.frame(x$metaData)
  idx <- which(tab$name == "InChIKey")
  return(tab$value[idx])
})

fls$author <- sapply(fls$metaData, function(x) {
  x<- as.data.frame(x) 
  ionization_row <- which(x$name == "author")
  return(x$value[ionization_row])
       })

fls <- fls %>%
  mutate(
    mz = lapply(spectrum, function(x) {
      if (!is.na(x)) as.numeric(str_extract_all(x, "[0-9\\.]+(?=:)")[[1]]) else NA
    }),
    intensity = lapply(spectrum, function(x) {
      if (!is.na(x)) as.numeric(str_extract_all(x, "(?<=:)[0-9\\.]+")[[1]]) else NA
    })
  )

#extracting class variable

fls$class <- sapply(fls$compound, function(x) {
  if (!"classification" %in% names(x) || is.null(x$classification)) {
    return(NA)  
  }
  class_df <- as.data.frame(x$classification)
  idx <- which(class_df$name == "class")
  if (length(idx) > 0) {
    return(class_df$value[idx])
  } else {
    return(NA)
  }
})
fls$subclass <- sapply(fls$compound, function(x) {
  if (is.null(x$classification)) {
    return(NA)  
  }
  class_df <- as.data.frame(x$classification)
  idx <- which(class_df$name == "subclass")
  if (length(idx) > 0) {
    return(class_df$value[idx])
  } else {
    return(NA)
  }
})
fls$superclass <- sapply(fls$compound, function(x) {
  if (is.null(x$classification)) {
    return(NA)  
  }
  class_df <- as.data.frame(x$classification)
  idx <- which(class_df$name == "superclass")
  if (length(idx) > 0) {
    return(class_df$value[idx])
  } else {
    return(NA)
  }
})
head(fls)
```

producing spectra object

```{r}
fls$polarity <- ifelse(fls$pol=="positive",1,0) %>%
  as.integer()
fls$precursorMz <- as.numeric(fls$precursorMz)
fls<- fls[which(sapply(fls$inchi, length)==1),]
allsps <- fls%>%
  dplyr::select(-"compound", "annotations", "metaData", "spectrum","pol") %>%
  Spectra(.)
allsps$inchi <- unlist(allsps$inchi)
allsps$id <- allsps$inchi

```

selecting the common spectra and making neutral accodring to inchi varaible of spectra. Considering fragemntation spectra from "[M-H]-" and "[M+H]+" adduct only.

```{r}
MS2neg <- allsps[allsps$polarity==0] 
MS2neg <- MS2neg[which(MS2neg$adduct=="[M-H]-")]
MS2pos <- allsps[allsps$polarity==1]
MS2pos <- MS2pos[which(MS2pos$adduct=="[M+H]+")]

comminchi <- intersect(MS2neg$inchi, MS2pos$inchi)  
MS2neg <- MS2neg[which(MS2neg$inchi %in% comminchi)]
MS2pos <- MS2pos[which(MS2pos$inchi %in% comminchi)]

#if duplicated, I keep the one with lowest entropy                 
lis <- c()
if(length(table(duplicated(MS2neg$inchi)))){
  d<- unique(MS2neg$inchi[which(duplicated(MS2neg$inchi))])
  for(i in 1:length(d)){
    e<-c()
    spx <- which(MS2neg$inchi ==d[i])
    idx <- spx[which.min(MS2neg$entropy[spx])]
    e <- spx[spx!=idx]
    lis<- c(e, lis)
  }
  lis <- unique(lis)
}
tokeep <-  setdiff(seq_along(MS2neg), lis)
MS2neg <- MS2neg[tokeep]

MS2pos <- MS2pos[which(MS2pos$inchi %in% comminchi)]
```

cleaning the spectra with more than 30 ions 

```{r}
MS2pos_cl <- cleaning_ref(MS2pos)
MS2neg_cl <- cleaning_ref(MS2neg)

hist(unlist(lapply(mz(MS2pos_cl), function(x) length(x))))
```

creating neutral pseudo-spectra for each compound that have negative and positive MS/MS spectra

```{r}
MS2neg_cl$inchi <- MS2neg_cl$id
MS2pos_cl$inchi <- MS2pos_cl$id

neutral_spectra<- Spectra()

for(i in 1:length(MS2pos_cl)){
  register(SerialParam())
  #positive
  fl <- MS2pos_cl
  akap <- fl[i]
    if(length(akap)>0){
    mzpos <- mz(akap)[[1]] 
    mzpos <- mzpos - 1.00728
    intpos <- intensity(akap)[[1]]
    intpos <- (intpos/max(intpos))*100
    df_pos <- data.frame(mz = mzpos, intensity = intpos)
    #negative
    fl <- MS2neg_cl
    akan <- fl[which(fl$inchi==akap$inchi)]
        if(length(akan)>0){
      mzneg <- mz(akan)[[1]] 
      mzneg <- mzneg + 1.00728
      intneg <- intensity(akan)[[1]]
      intneg <- (intneg/max(intneg))*100
      df_neg <- data.frame(mz = mzneg, intensity = intneg) 
          
      #both
      df <- bind_rows(df_neg, df_pos) %>%
        arrange(mz)
      df<- as.data.frame(process_data(df))
      totmz<- as.numeric(df$mz)
      totint <- as.numeric(df$intensity)
      
      prec<- mean(c(precursorMz(akan),precursorMz(akap)))
      
      spectrum<- data.frame(
        id= akap$inchi,
        mz = I(list(totmz)),
        intensity = I(list(totint)),
        precursorMz=prec
      )
      sps <- Spectra(spectrum)
      neutral_spectra <- c(neutral_spectra, sps)
    }
  }
  print(i)
}

#keeping the class and subclass varibales
neutral_spectra$class <- NA
for(i in 1:length(neutral_spectra)){
  neutral_spectra$class[i] <- MS2pos_cl$class[which(MS2pos_cl$inchi==neutral_spectra$id[i])]
}

neutral_spectra$subclass <- NA
for(i in 1:length(neutral_spectra)){
  neutral_spectra$subclass[i] <- MS2pos_cl$subclass[which(MS2pos_cl$inchi==neutral_spectra$id[i])]
}
```