---
title: "MNs_evaluation"
author: "Nicola Dalla Valle"
date: "2025-08-25"
output: html_document
---

```{r}
library(Spectra)
library(MsCoreUtils)
library(BiocParallel)
library(igraph)
library(dplyr)
library(tidyverse)
```

Code for Molecular Network building and their comparison according to structural metrics, as assortativity and modularity, based on both chemical classes and subclasses

assumning that we have produced the 3 polarities MS/MS datasets according to the .rmd "libraries_neutral_spectra_creation", we can proceed with the calcualtion of similarity matrices using GNPS score calcualtion

making matrices

```{r}
#neutral
matrix <- compareSpectra(MS2neu_cl, MAPFUN = joinPeaksGnps, tolerance = 0.01, 
                         FUN = MsCoreUtils::gnps)
rownames(matrix) <- MS2neu_cl$id
colnames(matrix) <- MS2neu_cl$id
#positive
matrix_pos <- compareSpectra(MS2pos_cl, MAPFUN = joinPeaksGnps, tolerance = 0.01, 
                             FUN = MsCoreUtils::gnps, BPPARAM = SnowParam(10))
rownames(matrix_pos) <- MS2pos_cl$id
colnames(matrix_pos) <- MS2pos_cl$id
#negative
matrix_neg <- compareSpectra(MS2neg_cl, MAPFUN = joinPeaksGnps, tolerance = 0.01, 
                             FUN = MsCoreUtils::gnps, BPPARAM = SnowParam(10))
rownames(matrix_neg) <- MS2neg_cl$id
colnames(matrix_neg) <- MS2neg_cl$id
```

considering only class with more than 10 spectra

```{r}
cltab <- as.data.frame(table(MS2pos_cl$class)) %>%
  filter(Freq>10)
filter_pos <- MS2pos_cl[which(MS2pos_cl$class %in% cltab$Var1)]
filtered<- data.frame(id=filter_pos$id,
                           class=filter_pos$class)
mtx <-matrix[which(rownames(matrix) %in% filtered$id),which(rownames(matrix) %in% filtered$id)]
mtx_pos <-matrix_pos[which(rownames(matrix_pos) %in% filtered$id),which(rownames(matrix_pos) %in% filtered$id)]
mtx_neg <-matrix_neg[which(rownames(matrix_neg) %in% filtered$id),which(rownames(matrix_neg) %in% filtered$id)]
```

calcualting assortativity and modualrity for stratified resampling of 80% of library spectra for 25 times in order to consider varaibility 
We are considering all range of similarity threshold, from 0 to 1.

```{r}
rep=25 #number of repetition

replicates <- seq(1, rep, by= 1)
reptab <- data.frame(n=replicates)
reptab$data <- vector(mode="list", length = nrow(reptab))

#making a weigthed sampling based on class
class_weights <- filtered %>%
  dplyr::count(class) %>%
  mutate(weight = round(n*0.8)) 

#loop for chemical similarity of connected nodes for each thr
for(q in 1:nrow(reptab)){
  #matirci 
  sp <- data.frame(cl=class_weights$class)
  sp$idn <- vector(mode="list", length=nrow(sp))
  for(v in 1:nrow(class_weights)){
    cl <- class_weights$class[v]
    idx <- MS2pos_cl$id[which(MS2pos_cl$class == cl)]
    idn <- sample(idx,class_weights$weight[v])
    sp$idn[[v]] <- idn
  }
  sp <- unlist(sp$idn)
  sneu <- MS2neu_cl[which(MS2neu_cl$id%in% sp)]
  sneg <- MS2neg_cl[which(MS2neg_cl$id%in% sp)]
  spos <-MS2pos_cl[which(MS2pos_cl$id%in% sp)]
  matrix2 <- mtx[which(rownames(mtx) %in% sneu$id),which(rownames(mtx) %in% sneu$id)]
  matrix_pos2 <- mtx_pos[which(rownames(mtx_pos) %in% spos$id),which(rownames(mtx_pos) %in% spos$id)]
  matrix_neg2 <- mtx_neg[which(rownames(mtx_neg) %in% sneg$id),which(rownames(mtx_neg) %in% sneg$id)]
  
  megatab<- data.frame(thr=seq(0,1,0.025),
                       Apos=NA,
                       Mpos=NA,
                       Aneu=NA,
                       Mneu=NA,
                       Aneg=NA,
                       Mneg=NA,
                       Apos_scl=NA,
                       Mpos_scl=NA,
                       Aneu_scl=NA,
                       Mneu_scl=NA,
                       Aneg_scl=NA,
                       Mneg_scl=NA,
                       Eneu=NA,
                       Epos=NA,
                       Eneg=NA)
  
  #calcolo similairtà per ogni fisrt degree parent
  
  for(k in 1:nrow(megatab)){
    thr<- megatab$thr[k]
    #Neutral graph
    all_nodes <- data.frame(
      id = sneu$id, 
      class=sneu$class,
      subclass=sneu$subclass
    )
    edges<- create_edge_list(matrix2, thr, sneu)
    megatab$Eneu[k]<- nrow(edges)
    graph<- graph_from_data_frame(d = edges, vertices = all_nodes, directed = FALSE)
    graph_neu<- graph
    #Positive graph
    all_nodes <- data.frame(
      name=spos$id,
      class=spos$class,
      subclass=spos$subclass
    )
    edges<- create_edge_list(matrix_pos2, thr, spos)
    megatab$Epos[k]<- nrow(edges)
    graph<- graph_from_data_frame(d = edges, vertices = all_nodes, directed = FALSE)
    graph_pos<- graph
    #Negative graph
    all_nodes <- data.frame(
      name=sneg$id,
      class=sneg$class,
      subclass=sneg$subclass
    )
    edges<- create_edge_list(matrix_neg2, thr, sneg)
    megatab$Eneg[k]<- nrow(edges)
    graph<- graph_from_data_frame(d = edges, vertices = all_nodes, directed = FALSE)
    graph_neg<- graph

    #class
    V(graph_neu)$class <-as.integer(as.factor(V(graph_neu)$class))
    V(graph_neu)$class[is.na(V(graph_neu)$class)]<-99
    megatab$Aneu[k]<-assortativity_nominal(graph_neu,V(graph_neu)$class)
    megatab$Mneu[k]<-modularity(graph_neu,membership=V(graph_neu)$class)
    #subclass
    V(graph_neu)$subclass <-as.integer(as.factor(V(graph_neu)$subclass))
    V(graph_neu)$subclass[is.na(V(graph_neu)$subclass)]<-99
    megatab$Aneu_scl[k]<-assortativity_nominal(graph_neu,V(graph_neu)$subclass)
    megatab$Mneu_scl[k]<-modularity(graph_neu,membership=V(graph_neu)$subclass)
    #graph pos 
    #class
    V(graph_pos)$class <-as.integer(as.factor(V(graph_pos)$class))
    V(graph_pos)$class[is.na(V(graph_pos)$class)]<-99
    megatab$Apos[k]<-assortativity_nominal(graph_pos,V(graph_pos)$class)
    megatab$Mpos[k]<-modularity(graph_pos,membership=V(graph_pos)$class)
    #subclass
    V(graph_pos)$subclass <-as.integer(as.factor(V(graph_pos)$subclass))
    V(graph_pos)$subclass[is.na(V(graph_pos)$subclass)]<-99
    megatab$Apos_scl[k]<-assortativity_nominal(graph_pos,V(graph_pos)$subclass)
    megatab$Mpos_scl[k]<-modularity(graph_pos,membership=V(graph_pos)$subclass)
    #graph neg
    #class
    V(graph_neg)$class <-as.integer(as.factor(V(graph_neg)$class))
    V(graph_neg)$class[is.na(V(graph_neg)$class)]<-99
    megatab$Aneg[k]<-assortativity_nominal(graph_neg,V(graph_neg)$class)
    megatab$Mneg[k]<-modularity(graph_neg,membership=V(graph_neg)$class)
    #subclass
    V(graph_neg)$subclass <-as.integer(as.factor(V(graph_neg)$subclass))
    V(graph_neg)$subclass[is.na(V(graph_neg)$subclass)]<-99
    megatab$Aneg_scl[k]<-assortativity_nominal(graph_neg,V(graph_neg)$subclass)
    megatab$Mneg_scl[k]<-modularity(graph_neg,membership=V(graph_neg)$subclass)
  }
  
  reptab$data[[q]]<- megatab
  
}

```
