---
title: "neighbors_subclass_matches"
author: "Nicola Dalla Valle"
date: "2025-08-25"
output: html_document
---

```{r}
create_edge_list <- function(similarity_matrix, thr, spectra) {
  non_zero_indices <- which(similarity_matrix > thr, arr.ind = TRUE)
  if (length(non_zero_indices) > 0 && is.null(dim(non_zero_indices))) {
    non_zero_indices <- matrix(non_zero_indices, ncol = 2, byrow = TRUE)
  }
  if (length(non_zero_indices) == 0) {
    return(data.frame(from = character(), to = character(), weight = numeric()))
  }
  non_zero_indices <- non_zero_indices[non_zero_indices[, 1] < non_zero_indices[, 2], , drop = FALSE]
  if (nrow(non_zero_indices) == 0) {
    return(data.frame(from = character(), to = character(), weight = numeric()))
  }
  # Create an edge list from these indices
  edges <- data.frame(
    from = spectra$id[non_zero_indices[, 1]],
    to = spectra$id[non_zero_indices[, 2]],
    weight = similarity_matrix[non_zero_indices]
  )
  
  return(edges)
}
```

assuming thath we have our dataset divided by polarity (as described in "libraries_neutral_spectra_creation.Rmd") and we have calcualted our matrices of spectra similarity, we can porceed we the evaluation of ability to predict the subcless of a putative "unknown" node in MNs acording to a majority vote considering their neighburs. Calculation was made considering up to the top 10 neighbors according to spectral similarity applied to a zero MN threshold.

Creation of MNs for all the 3 polarties

```{r}
#handling NAs values, if present
sub <- MS2neu_cl[!is.na(MS2neu_cl$subclass)]
subn <- MS2neg_cl[!is.na(MS2neg_cl$subclass)]
subp <- MS2pos_cl[!is.na(MS2pos_cl$subclass)]

matrix2 <-matrix[which(rownames(matrix) %in% sub$id),which(rownames(matrix) %in% sub$id)]
matrix_pos2 <-matrix_pos[which(rownames(matrix_pos) %in% subp$id),which(rownames(matrix_pos) %in% subp$id)]
matrix_neg2 <-matrix_neg[which(rownames(matrix_neg) %in% subn$id),which(rownames(matrix_neg) %in% subn$id)]

thr<- 0.0
#neutral
all_nodes <- data.frame(
  id = sub$id, 
  class=subp$subclass
)
edges<- create_edge_list(matrix2, thr, sub)
gneu <- graph_from_data_frame(d = edges, vertices = all_nodes, directed = FALSE)

#pos
all_nodes_p <- data.frame(
  id = subp$inchi,
  class=subp$subclass
)
edges<- create_edge_list(matrix_pos2, thr, subp)
gpos <- graph_from_data_frame(d = edges, vertices = all_nodes_p, directed = FALSE)

#neg
all_nodes_n <- data.frame(
  id = subn$inchi,
  class=subn$subclass
)
edges<- create_edge_list(matrix_neg2, thr, subn)
gneg <- graph_from_data_frame(d = edges, vertices = all_nodes_n, directed = FALSE)
```

calcualtion of correct chemical subclass prediction according to majority vote
25 repetition to estimate variability, random selection of 50 compounds within each network where mean of correct prediction were considered.

```{r}
gigatab<- data.frame(rep=seq(1,25,1))
gigatab$data <- vector(mode="list", length=nrow(gigatab))

for(w in 1:nrow(gigatab)){
  seq <- seq(1,10,1)
  maintab<- data.frame(n=seq,
                       neu=NA,
                       pos=NA,
                       neg=NA)
  maintab$data <- vector(mode="list", length=nrow(maintab))
  baba<- MS2neu_cl[!is.na(MS2neu_cl$subclass)]
  
  ink<- sample(baba$id,50)
  
  for(j in 1:nrow(maintab)){
    ntab <- data.frame(n=ink,
                       neu=NA,
                       pos=NA,
                       neg=NA,
                       real_cl=NA)
    for(k in 1:nrow(ntab)){
      name <- ntab$n[k]
      real_cl <- MS2neu_cl$subclass[which(MS2neu_cl$id==name)]
      ntab$real_cl[k]<- real_cl
      #neutral
      edges <- E(gneu)[incident(gneu, name)]
      edges_df <- data.frame(
        from = ends(gneu, edges)[, 1],
        to = ends(gneu, edges)[, 2],
        weight = E(gneu)[edges]$weight
      ) %>%
        arrange(desc(weight))
      edges_df <- edges_df[1:j,]
      edges_df <- edges_df[!is.na(edges_df$weight),]
      edges_df$conn<-NA
      edges_df$conn[which(edges_df$from!=name)] <- edges_df$from[which(edges_df$from!=name)]
      edges_df$conn[which(edges_df$to!=name)] <- edges_df$to[which(edges_df$to!=name)]
      edges_df$class <- sapply(edges_df$conn,function(x) MS2neu_cl$subclass[which(MS2neu_cl$id==x)])
      if(length(edges_df$class )>0){
        stimated_cl <-  names(which.max(table(edges_df$class)))
        ntab$neu[k] <- stimated_cl
        weighted_votes <- aggregate(weight ~ class, data = edges_df, sum)
        stimated_cl <- weighted_votes$class[which.max(weighted_votes$weight)]
        ntab$wneu[k] <- stimated_cl
      }
      #positive
      edges <- E(gpos)[incident(gpos, name)]
      edges_df <- data.frame(
        from = ends(gpos, edges)[, 1],
        to = ends(gpos, edges)[, 2],
        weight = E(gpos)[edges]$weight
      ) %>%
        arrange(desc(weight))
      edges_df <- edges_df[1:j,]
      edges_df <- edges_df[!is.na(edges_df$weight),]
      edges_df$conn<-NA
      edges_df$conn[which(edges_df$from!=name)] <- edges_df$from[which(edges_df$from!=name)]
      edges_df$conn[which(edges_df$to!=name)] <- edges_df$to[which(edges_df$to!=name)]
      edges_df$class <- sapply(edges_df$conn,function(x) MS2neu_cl$subclass[which(MS2neu_cl$id==x)])
      if(length(edges_df$class )>0){
        stimated_cl <-  names(which.max(table(edges_df$class)))
        ntab$pos[k] <- stimated_cl
        weighted_votes <- aggregate(weight ~ class, data = edges_df, sum)
        stimated_cl <- weighted_votes$class[which.max(weighted_votes$weight)]
        ntab$wpos[k] <- stimated_cl
      }
      
      #negative
      edges <- E(gneg)[incident(gneg, name)]
      edges_df <- data.frame(
        from = ends(gneg, edges)[, 1],
        to = ends(gneg, edges)[, 2],
        weight = E(gneg)[edges]$weight
      ) %>%
        arrange(desc(weight))
      edges_df <- edges_df[1:j,]
      edges_df <- edges_df[!is.na(edges_df$weight),]
      edges_df$conn<-NA
      edges_df$conn[which(edges_df$from!=name)] <- edges_df$from[which(edges_df$from!=name)]
      edges_df$conn[which(edges_df$to!=name)] <- edges_df$to[which(edges_df$to!=name)]
      edges_df$class <- sapply(edges_df$conn,function(x) MS2neu_cl$subclass[which(MS2neu_cl$id==x)])
      if(length(edges_df$class )>0){
        stimated_cl <-  names(which.max(table(edges_df$class)))
        ntab$neg[k] <- stimated_cl
        weighted_votes <- aggregate(weight ~ class, data = edges_df, sum)
        stimated_cl <- weighted_votes$class[which.max(weighted_votes$weight)]
        ntab$wneg[k] <- stimated_cl
      }
    }
    maintab$data[[j]] <- ntab
    maintab$neu[j]<-length(which(ntab$neu==ntab$real_cl))
    maintab$pos[j]<-length(which(ntab$pos==ntab$real_cl))
    maintab$neg[j]<- length(which(ntab$neg==ntab$real_cl))
    
  }
  print(w)
  gigatab$data[[w]]<- maintab
}

all_res <- gigatab %>%
  mutate(rep_id = row_number()) %>%
  dplyr::select(rep_id, data) %>%
  unnest(data) 

summary_res <- all_res %>%
  group_by(n) %>%
  summarise(
    neu_mean = mean(neu, na.rm=TRUE),
    pos_mean = mean(pos, na.rm=TRUE),
    neg_mean = mean(neg, na.rm=TRUE),
    neu_sd = sd(neu, na.rm=TRUE),
    pos_sd = sd(pos, na.rm=TRUE),
    neg_sd = sd(neg, na.rm=TRUE),
    .groups="drop"
  )

```





