---
title: "alt_allMNs"
output: html_document
date: "2026-04-08"
---

```{r}
library(xcms)
library(Spectra)
library(MsCoreUtils)
library(tidyverse)
library(igraph)
library(visNetwork)
library(MASS)

cleaning_ref<- function(ms2){
  ms2_tb <- as.data.frame(
    spectraData(ms2) 
  ) %>%
  mutate(ms2_raw = map2(mz(ms2), intensity(ms2), ~{
    if (length(.x) == 0 || length(.y) == 0)
      return(data.frame(mz = numeric(0), intensity = numeric(0)))
    data.frame(mz = as.numeric(.x), intensity = as.numeric(.y))
  })) %>%
    mutate(ms2_RLM = purrr::map(ms2_raw, function(t){
         if (is.null(t) || nrow(t) == 0 || !"intensity" %in% names(t))
        return(tibble(mz = numeric(0), intensity = numeric(0)))
      t <- t %>%
        filter(intensity>0) %>%
        arrange(intensity) %>%
        mutate(order = seq(nrow(.)))
      p75 <- t %>%
        pull(intensity) %>% 
        quantile(0.75)
      u <- t %>%
        filter(intensity < p75)
      if(nrow(u)>3){
        mdl <- rlm(data = u, log10(intensity) ~ order)
        t_filt<- t %>% 
          mutate(pred = predict(mdl, t)) %>%
          mutate(res = abs(pred - log10(intensity))) %>%
          filter(res > sd(residuals(mdl))*3) %>%
          arrange(mz)
        if (nrow(t_filt) > 0) {
          t<- t_filt
        } else {
          t<- t %>% arrange(mz)
        }
      } else {
        return(t %>% arrange(mz))  # Not enough points to model
      }
    }))
  ms2_cl <- ms2_tb %>%
    mutate(
      mz = purrr::map(ms2_RLM, ~ .x$mz),
      intensity = purrr::map(ms2_RLM, ~ .x$intensity)
    ) %>%
    filter(map_lgl(mz, ~ length(.x) > 0)) %>%
    as.data.frame() %>%
    Spectra()
  return(ms2_cl)
}

#loading data from neutral pseudo-spectra pairing and creation
load("rdata/alt_msms.RData")
```

The aim is to see changes between plots before and after the subsetting realtive to neutral pseudo-spectra subset selection.
First we need to compute matrices for entire networks, considering original feature detection obteined after feature creation in xcms processs.

#Negative

```{r}
register(SerialParam())
spsneg<- cleaning_ref(spsneg) #all dataset of feature having a MS/MS spectra
mat_neg_all <- compareSpectra(spsneg, MAPFUN = joinPeaksGnps, tolerance = 0.01, FUN = MsCoreUtils::gnps)
  colnames(mat_neg_all)<- rownames(mat_neg_all) <- spsneg$feature_id

save(mat_neg_all, file="rdata/alt_all_MNs.RData") 
```

Comparing the full network against subsetting used for neutral pseudo-spectra creation
Comparison with random subsets mantaining the same size ids done to evaluate the artifact of size decrease (100 repetitions).

```{r}
threshold<- 0.7

#total
adj <- mat_neg_all
adj[adj < threshold] <- 0
diag(adj) <- 0
gt <- graph_from_adjacency_matrix(adj, mode = "undirected", weighted = TRUE, diag = FALSE)

comp <-  cluster_louvain(gt)
gt$comm <- comp$membership

raw_mod <- modularity(gt, gt$comm)
raw_ass<- assortativity(gt, gt$comm)

#NMN subsetting
adj <- matrix_neg
adj[adj < threshold] <- 0
diag(adj) <- 0
g <- graph_from_adjacency_matrix(adj, mode = "undirected", weighted = TRUE, diag = FALSE)

comp <-  cluster_louvain(g)
g$comm <- comp$membership

filt_mod<- modularity(g, g$comm)
filt_ass<- assortativity(g, g$comm)

#random subsetting
#keeping the number equal to NMN subset
mods <- c()
asss <- c()
for(i in 1:100){
  idx <- sample(1:nrow(mat_neg_all), length(MS2neu))
adj <- mat_neg_all[idx,idx]
adj[adj < threshold] <- 0
diag(adj) <- 0
gx <- graph_from_adjacency_matrix(adj, mode = "undirected", weighted = TRUE, diag = FALSE)

comp <-  cluster_louvain(gx)
gx$comm <- comp$membership
mods<- c(modularity(gx, gx$comm), mods)
asss<- c(assortativity(gx, gx$comm), asss)
}

df_mod <- data.frame(
  value = mods,
  metric = "Modularity"
)
df_ass <- data.frame(
  value = asss,
  metric = "Assortativity"
)
df <- rbind(df_mod, df_ass)

refs <- data.frame(
  metric = c("Modularity", "Modularity", "Assortativity", "Assortativity"),
  type = c("Raw", "Filtered", "Raw", "Filtered"),
  value = c(raw_mod, filt_mod, raw_ass, filt_ass)
)

neg <- ggplot(df, aes(x = value)) +
  geom_histogram(bins = 30, fill = "grey80", color = "black") +
  
  geom_vline(data = refs,
             aes(xintercept = value, color = type, linetype = type),
             size = 1) +
  
  facet_wrap(~metric, ncol = 1, scales = "free") +
  
scale_color_manual(values = c(
  "Raw" = "#1B9E77",      # teal
  "Filtered" = "#D95F02"  # burnt orange
)) +
  labs(x = "Value", y = "Frequency") +
  
  theme_minimal() +
  theme(legend.position = "bottom",       
        legend.title = element_blank())
neg
```

#Positive

```{r}
register(SerialParam())
spspos<- cleaning_ref(spspos) #all dataset of feature having a MS/MS spectra
mat_pos_all <- compareSpectra(spspos, MAPFUN = joinPeaksGnps, tolerance = 0.01, FUN = MsCoreUtils::gnps)
  colnames(mat_pos_all)<- rownames(mat_pos_all) <- tmppos$feature_id

save(mat_neg_all,mat_pos_all, file="rdata/alt_all_MNs.RData") 
 
```

Comparing the full network against subsetting used for neutral pseudo-spectra creation
Comparison with random subsets mantaining the same size ids done to evaluate the artifact of size decrease (100 repetitions).

```{r}
threshold<- 0.7

#total
adj <- mat_pos_all
adj[adj < threshold] <- 0
diag(adj) <- 0
gt <- graph_from_adjacency_matrix(adj, mode = "undirected", weighted = TRUE, diag = FALSE)

comp <-  cluster_louvain(gt)
gt$comm <- comp$membership

raw_mod <- modularity(gt, gt$comm)
raw_ass<- assortativity(gt, gt$comm)

#NMN subsetting
adj <- matrix_neg
adj[adj < threshold] <- 0
diag(adj) <- 0
g <- graph_from_adjacency_matrix(adj, mode = "undirected", weighted = TRUE, diag = FALSE)

comp <-  cluster_louvain(g)
g$comm <- comp$membership

filt_mod<- modularity(g, g$comm)
filt_ass<- assortativity(g, g$comm)

#random subsetting
#keeping the number equal to NMN subset
mods <- c()
asss <- c()
for(i in 1:100){
  idx <- sample(1:nrow(mat_pos_all), length(MS2neu))
adj <- mat_pos_all[idx,idx]
adj[adj < threshold] <- 0
diag(adj) <- 0
gx <- graph_from_adjacency_matrix(adj, mode = "undirected", weighted = TRUE, diag = FALSE)

comp <-  cluster_louvain(gx)
gx$comm <- comp$membership
mods<- c(modularity(gx, gx$comm), mods)
asss<- c(assortativity(gx, gx$comm), asss)
}

df_mod <- data.frame(
  value = mods,
  metric = "Modularity"
)
df_ass <- data.frame(
  value = asss,
  metric = "Assortativity"
)
df <- rbind(df_mod, df_ass)

refs <- data.frame(
  metric = c("Modularity", "Modularity", "Assortativity", "Assortativity"),
  type = c("Raw", "Filtered", "Raw", "Filtered"),
  value = c(raw_mod, filt_mod, raw_ass, filt_ass)
)

pos<- ggplot(df, aes(x = value)) +
  geom_histogram(bins = 30, fill = "grey80", color = "black") +
  
  geom_vline(data = refs,
             aes(xintercept = value, color = type, linetype = type),
             size = 1) +
  
  facet_wrap(~metric, ncol = 1, scales = "free") +
  
scale_color_manual(values = c(
  "Raw" = "#1B9E77",      # teal
  "Filtered" = "#D95F02"  # burnt orange
)) +
  labs(x = "Value", y = "Frequency") +
  
  theme_minimal() +
  theme(legend.position = "none",
        legend.title = element_blank())
pos
```
